---
title: "Enrichment Analysis"
output: word_document
---

Enrichment analysis given clusters

# packages
```{r echo = FALSE, message = FALSE}


```

# functions

enrichment score
in: x array (unordered)
    in or out matrix (ions_inpath)
    pathway of interest
    plot?
    cluster description if applicable
    jumblings (how many times) (default 1000)
out: ret
      ret$x_array = ion values (ordered)
      ret$e_score = e score at that x value
      ret$Q = enrichment score
      ret$Q_extreme = maximum possible enrichment score given that all "on" ions are at the beginning of the sequence
      ret$probability = percent of randomly created enrichment scores which fall below Q
```{r echo = FALSE, message = FALSE}

escore <- function(data) {
  sum_in <- sum((data[data$in_path == 1,]$x_array)^2)
  sum_out <- sum((data[data$in_path == 0,]$x_array)^2)
  
  w <- sum_out / sum_in
  data$e_score = 0
  amp <- 0

  for (ion in 1:nrow(data)) {
    if (data$in_path[ion] == 1) {
      amp <- amp + (w * (data$x_array[ion])^2)
    } else if (data$in_path[ion] == 0) {
      amp <- amp - (data$x_array[ion])^2
    }
    data$e_score[ion] <- amp
  }  
  
  if (abs(min(data$e_score)) > max(data$e_score)) {es <- min(data$e_score)} else {es <- max(data$e_score)}
  res <- list("data" = data,
              "es" = es)
  return(res)
}



calc_ES <- function(x_array, pathway, ions_inpath, plot, cluster_description, jumblings = 1000) {
  
  in_path <- ions_inpath[,grep(paste0("^", pathway, "$"), names(ions_inpath))]
  if (!is.integer(in_path)) {
    print("pathway not found")
  } else {
    data <- data.frame(x_array, in_path)
    data <- data[order(data$x_array),]
    
    copy <- data
    escore_data <- escore(data)
    data <- escore_data$data
    es <- escore_data$es
    data$x_step <- seq(2:(nrow(data)+1))
    
    if (plot == TRUE) {
      par(mar = c(5,5,2,5))
      plot(c(1,data$x_step), c(0,data$e_score), type = "l", ylab = "enrichment score", xlab = "index")
      abline(h = 0)
      par(new = T)
      plot(data$x_step, (data$x_array^2), xlab = NA, ylab = NA, axes = F, type = "l", col = "red")
      axis(side = 4)
      mtext(side = 4, line = 3, "squared ion means")
      title(paste0("pathway: ", pathway, " | cluster: ", cluster_description))
    }
    
  
    # jumble weights 1000 times and store min and max values
    j_es <- c()
    
    for (i in 1:jumblings) {
      jumbled <- copy
      jumbled$in_path <- sample(copy$in_path, size = length(copy$in_path), replace = FALSE)
      escore_jumbled<- escore(jumbled)
      jumbled <- escore_jumbled$data
      j_es <- c(j_es, escore_jumbled$es)
    }
    
    extreme <- copy
    extreme$in_path <- c(rep(1, sum(extreme$in_path)), rep(0, length(extreme$in_path) - sum(extreme$in_path)))
    extreme_val <- escore(extreme)$es
    
    if (es < quantile(j_es, .05)) {pmin = TRUE} else {pmin = FALSE}
    if (es > quantile(j_es, .95)) {pmax = TRUE} else {pmax = FALSE} 
    
    p <- sum(j_es < es)/length(j_es)
    
    ret <- list(x_array = data$x_array, 
                e_score = data$e_score, 
                Q = es,
                Q_extreme = extreme_val,
                probability = p
                )
    
    return(ret)
  }
}


```

Create enrichment score summary for each method
in: list of pathways (names) of interest
    coordinates and clusters (results of cluster_data)
    metabolite data
    ions_inpath
out: summary of pathway, cluster, Q, Q_extreme, # sd away from mean
```{r echo = FALSE, method = FALSE}

es_summary <- function(pathways, clusters, inpath_matrix, ion_data) {
  summary <- data.frame(matrix(ncol = 5, nrow = length(pathways)*length(unique(clusters$clusters))))
  names(summary) <- c("pathway", "cluster", "Q", "Q_extreme", "probability")
  i <- 1
  for (p in pathways) {
    for (c in unique(clusters$clusters)){
      df <- ion_data[which(clusters$clusters == c),]
      res <- calc_ES(x_array = colMeans(df), pathway = p, ions_inpath = inpath_matrix, plot = TRUE, cluster_description = c, jumblings = 1000)
      summary$pathway[i] <- p
      summary$cluster[i] <- c
      summary$Q[i] <- res$Q
      summary$Q_extreme[i] <- res$Q_extreme
      summary$probability[i] <- res$probability
      i <- i + 1
    }
  }
  return(summary)
}

```

Find "important" pathways; those with the most ions in them
in: binary matrix of ions by pathway (from prev)
    n pathways to find (default 5)
out: list  top n pathways
```{r echo = FALSE, message = FALSE}
find_important_pathways <- function(ion_inpath, n = 5){
  pathway_summary <-  data.frame("pathway" = names(ion_inpath), "total_ions" = colSums(ion_inpath))
  pathway_summary <- pathway_summary[order(pathway_summary$total_ions, decreasing = TRUE),]
  return(as.character(pathway_summary$pathway[1:n]))
}

```



# load data

```{r echo = FALSE, message = FALSE}

ions_inpath <- read.csv("../data/ion_inpath.csv")

pca_train_clusters <- read.csv("../data/pca_train_clusters.csv")
pca_val_clusters <- read.csv("../data/pca_val_clusters.csv")

tsne_train_clusters <- read.csv("../data/tsne_train_clusters.csv")
tsne_val_clusters <- read.csv("../data/tsne_val_clusters.csv")

som_train_clusters <- read.csv("../data/som_train_clusters.csv")
# som val clusters have not been made yet
# som_val_clusters <- read.csv("../data/som_val_clusters.csv")

untar <- read.csv("../data/zscore_untargeted_annotatedions.csv")
train <- untar[untar$group <= 2, -c(1:3)]
val <- untar[untar$group == 3, -c(1:3)] 

```

# Find  pathways for enrichment analysis (top = pathways with most ions in them)
```{r echo = FALSE, message = FALSE}

top_pathways <- find_important_pathways(ions_inpath, n = 20)

```

# PCA enrichment analysis
```{r echo = FALSE, message = FALSE}

pca_es_summary <- es_summary(pathways = top_pathways, clusters = pca_train_clusters, inpath_matrix = ions_inpath, ion_data = train)
write.csv(pca_es_summary, "../data/pca_es_summary.csv", row.names = FALSE)


pca_val_es_summary <- es_summary(pathways = top_pathways, clusters = pca_val_clusters, inpath_matrix = ions_inpath, ion_data = val)
write.csv(pca_val_es_summary, "../data/pca_val_es_summary.csv", row.names = FALSE)

```


# TSNE enrichment analysis
```{r echo = FALSE, message = FALSE}

tsne_es_summary <- es_summary(pathways = top_pathways, clusters = tsne_train_clusters, inpath_matrix = ions_inpath, ion_data = train)
write.csv(tsne_es_summary, "../data/tsne_es_summary.csv", row.names = FALSE)

tsne_val_es_summary <- es_summary(pathways = top_pathways, clusters = tsne_val_clusters, inpath_matrix = ions_inpath, ion_data = val)
write.csv(tsne_val_es_summary, "../data/tsne_val_es_summary.csv", row.names = FALSE)

```

# SOM enrichment analysis
```{r echo = FALSE, message = FALSE}

som_es_summary <- es_summary(pathways = top_pathways, clusters = som_train_clusters, inpath_matrix = ions_inpath, ion_data = train)
write.csv(som_es_summary, "../data/som_es_summary.csv", row.names = FALSE)

```























