---
title: "Methods for Dimension Reduction Analysis"
output: word_document
---
# packages
```{r echo = FALSE, message = FALSE}
library(DescTools)

```



hierarchichal clustering
in: (x,y) coordinates
out: clusters column added to initial df
     plots x,y coordinates and clusters
```{r echo = FALSE, message = FALSE}


plot_cluster=function(data, var_cluster, point_size) {
  ggplot(data, aes_string(x="V1", y="V2", color=var_cluster)) +
  geom_point(size=point_size) +
  guides(colour=guide_legend(override.aes=list(size=6))) +
  xlab("") + ylab("") +
  ggtitle("") +
  theme_light(base_size=20) +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        legend.direction = "horizontal", 
        legend.position = "bottom",
        legend.box = "horizontal")
}



cluster_data <- function(coords, point_size = 10, h) {
  names(coords) = c("V1", "V2")
  fit_cluster_hierarchical=hclust(dist(scale(coords)))
  plot(fit_cluster_hierarchical)
  coords$clusters = factor(cutree(fit_cluster_hierarchical, h=h))
  plot(plot_cluster(coords, coords$clusters, point_size))
  return(coords)
}


```


enrichment score
in: x array (unordered)
    in or out matrix (ions_inpath)
    pathway of interest
    plot?
out: res
      res$x_array = ion values (ordered)
      res$e_score = e score at that x value
      res$pmin = p min (p value comparing minimum to jumbled distribution) >> currently TRUE if value above 95% interval
      res$pmax = p max (p value comparing maximum to jumbled distribution) >> or belot 5% interval
```{r echo = FALSE, message = FALSE}

escore <- function(data) {
  sum_in <- sum((data[data$in_path == 1,]$x_array)^2)
  sum_out <- sum((data[data$in_path == 0,]$x_array)^2)
  
  w <- sum_out / sum_in
  data$e_score = 0
  amp <- 0

  for (ion in 1:nrow(data)) {
    if (data$in_path[ion] == 1) {
      amp <- amp + (w * (data$x_array[ion])^2)
    } else if (data$in_path[ion] == 0) {
      amp <- amp - (data$x_array[ion])^2
    }
    data$e_score[ion] <- amp
  }  
  
  if (abs(min(data$e_score)) > max(data$e_score)) {es <- min(data$e_score)} else {es <- max(data$e_score)}
  res <- list("data" = data,
              "es" = es)
  return(res)
}



calc_ES <- function(x_array, pathway, ions_inpath, plot, cluster_description) {
  
  in_path <- ions_inpath[,grep(paste0(pathway, "$"), names(ions_inpath))]
  if (!is.integer(in_path)) {
    print("pathway not found")
  } else {
    data <- data.frame(x_array, in_path)
    data <- data[order(data$x_array),]
    
    copy <- data
    escore_data <- escore(data)
    data <- escore_data$data
    es <- escore_data$es
    data$x_step <- seq(2:(nrow(data)+1))
    
    if (plot == TRUE) {
      par(mar = c(5,5,2,5))
      plot(c(1,data$x_step), c(0,data$e_score), type = "l", ylab = "enrichment score", xlab = "index")
      abline(h = 0)
      par(new = T)
      plot(data$x_step, (data$x_array^2), xlab = NA, ylab = NA, axes = F, type = "l", col = "red")
      axis(side = 4)
      mtext(side = 4, line = 3, "squared ion means")
      title(paste0("pathway: ", pathway, " | cluster: ", cluster_description))
    }
    
  
    # jumble weights 1000 times and store min and max values
    j_es <- c()
    
    for (i in 1:1000) {
      jumbled <- copy
      jumbled$in_path <- sample(copy$in_path, size = length(copy$in_path), replace = FALSE)
      escore_jumbled<- escore(jumbled)
      jumbled <- escore_jumbled$data
      j_es <- c(j_es, escore_jumbled$es)
    }
    
    extreme <- copy
    extreme$in_path <- c(rep(1, sum(extreme$in_path)), rep(0, length(extreme$in_path) - sum(extreme$in_path)))
    extreme_val <- escore(extreme)$es
    
    if (es < quantile(j_es, .05)) {pmin = TRUE} else {pmin = FALSE}
    if (es > quantile(j_es, .95)) {pmax = TRUE} else {pmax = FALSE} 
    
    p <- sum(abs(j_es) < es)/length(j_es)
    
    ret <- list(x_array = data$x_array, 
                e_score = data$e_score, 
                Q = es,
                Q_extreme = extreme_val,
                probability = p
                )
    
    return(ret)
  }

}


```
     

# Generate Ion-Pathway information for enrichment analysis
Before calculating enrichment score: make dataframe of ions and which pathways they are in
in: annotated ions
out: writes dataframe of ions and which pathways they are in as pathway_df.csv
```{r echo = FALSE, message = FALSE}
#library(BiocManager)
#BiocManager::install("hmdbQuery")

find_ions_in_pathways_nopunct <- function(ann) {
  library(hmdbQuery)
  
  # load annotated ions
  ann <- read.csv("../data/CRIC_annotation_1mD_neg.csv")
  
  pathway_df <- data.frame("init" = c(1:(nrow(ann)*100)),
                           "metabolite" = 0, 
                           "ion" = 0,
                           "HMDBID" = 0,
                           "pathway" = 0)
  pathway_df <- pathway_df[-1]
  line <- 1
  
  metabolites <- c()
  
  # for every metabolite in ann (i) 
  for (i in 1:nrow(ann)) { 
    met_IDs <- unlist(strsplit(as.character(ann$'ï..id'[i]), '; '))
    met_IDs <- met_IDs[grepl("HMDB", met_IDs)]
    met_IDs <- sub("HMDB", "HMDB00", met_IDs)
    metabolites <- c(metabolites, gsub("[[:punct:]]", "", ann$name[i]))
    
    for (id in met_IDs) {
      
      entry <- tryCatch(HmdbEntry(prefix = "http://www.hmdb.ca/metabolites/", id = id), error = function(e) entry <- NA)
      
      if (!is.na(entry)) {
        pathways <- store(entry)$biological_properties$pathways
        
        if (pathways != "\n    ") {
          prev_paths <- c()
          
          for(path in 1:length(pathways)){
            if (!gsub(" ", "", unlist(strsplit(as.character(pathways[path]$pathway$name), "[[:punct:]]"))[1]) %in% prev_paths){
              pathway_df$metabolite[line] <- gsub("[[:punct:]]", "", ann$name[i])
              pathway_df$ion[line] <- ann$ion[i]
              pathway_df$HMDBID[line] <- id
              pathway_df$pathway[line] <- gsub(" ", "", unlist(strsplit(as.character(pathways[path]$pathway$name), "[[:punct:]]"))[1])
              line <- line + 1
              prev_paths <- c(prev_paths, gsub(" ", "", unlist(strsplit(as.character(pathways[path]$pathway$name), "[[:punct:]]"))[1]))
            }
          }
        }
      }
    }
    print(paste0("met ", i, ": ", as.character(ann$name[i])))
  }
  
  pathway_df <- pathway_df[pathway_df$metabolite != 0,]
  
  write.csv(pathway_df, "../data/pathway_df_nopunct.csv", row.names = FALSE)
}

```

```{r echo = FALSE, message = FALSE}
#library(BiocManager)
#BiocManager::install("hmdbQuery")

find_ions_in_pathways <- function(ann) {
  library(hmdbQuery)
  
  # load annotated ions
  ann <- read.csv("../data/CRIC_annotation_1mD_neg.csv")
  
  pathway_df <- data.frame("init" = c(1:(nrow(ann)*100)),
                           "metabolite" = 0, 
                           "ion" = 0,
                           "HMDBID" = 0,
                           "pathway" = 0)
  pathway_df <- pathway_df[-1]
  line <- 1
  
  metabolites <- c()
  
  # for every metabolite in ann (i) 
  for (i in 1:nrow(ann)) { 
    met_IDs <- unlist(strsplit(as.character(ann$'ï..id'[i]), '; '))
    met_IDs <- met_IDs[grepl("HMDB", met_IDs)]
    met_IDs <- sub("HMDB", "HMDB00", met_IDs)
    metabolites <- c(metabolites, gsub("[[:punct:]]", "", ann$name[i]))
    
    for (id in met_IDs) {
      
      entry <- tryCatch(HmdbEntry(prefix = "http://www.hmdb.ca/metabolites/", id = id), error = function(e) entry <- NA)
      
      if (!is.na(entry)) {
        pathways <- store(entry)$biological_properties$pathways
        
        if (pathways != "\n    ") {
          prev_paths <- c()
          
          for(path in 1:length(pathways)){
            if (!gsub(" ", "", unlist(as.character(pathways[path]$pathway$name))) %in% prev_paths){
              pathway_df$metabolite[line] <- gsub("[[:punct:]]", "", ann$name[i])
              pathway_df$ion[line] <- ann$ion[i]
              pathway_df$HMDBID[line] <- id
              pathway_df$pathway[line] <- gsub(" ", "", unlist(as.character(pathways[path]$pathway$name) ))
              line <- line + 1
              prev_paths <- c(prev_paths, gsub(" ", "", unlist(as.character(pathways[path]$pathway$name))))
            }
          }
        }
      }
    }
    print(paste0("met ", i, ": ", as.character(ann$name[i])))
  }
  
  pathway_df <- pathway_df[pathway_df$metabolite != 0,]
  
  return(pathway_df)
  
  write.csv(pathway_df, "../data/pathway_df_punct.csv", row.names = FALSE)
}

```

Re format pathway_df as binary matrix
in: dataframe of ions and which pathways they are in (from prev)
out: writes binary matrix of ions by pathway as ion_inpath.csv
```{r echo = FALSE, message = FALSE}
binary_pathways <- function(pathway_df, name) {
  ion_inpath <- as.data.frame(matrix(nrow = length(unique(ann$ion)), ncol = length(unique(pathway_df$pathway)), 0))
  rownames(ion_inpath) <- unique(ann$ion)
  colnames(ion_inpath) <- unique(pathway_df$pathway)
  
  
  for (path in 1:ncol(ion_inpath)){
    ion_inpath[[path]] <- as.numeric(c(rownames(ion_inpath)) %in% unique(pathway_df[pathway_df$pathway == names(ion_inpath)[path],]$ion))
  }
  
  write.csv(ion_inpath, paste0("../data/ion_inpath_", name, ".csv"), row.names = FALSE)
}

```
     
Find "important" pathways; those with the most ions in them
in: binary matrix of ions by pathway (from prev)
```{r echo = FALSE, message = FALSE}
find_important_pathways <- function(ion_inpath, n = 5){
  pathway_summary <-  data.frame("pathway" = names(ion_inpath), "total_ions" = colSums(ion_inpath))
  pathway_summary <- pathway_summary[order(pathway_summary$total_ions, decreasing = TRUE),]
  return(as.character(pathway_summary$pathway[1:n]))
}

```
     
# tSNE
test several options and  choose optimal perplexity
in: data, perplexities
out: plots
```{r echo = FALSE, message = FALSE}
library(Rtsne)
library(ggplot2)
find_optimal_p_tSNE <- function(perplexities, data) {
  for (p in perplexities) {
    tsne_out <- Rtsne(as.matrix(train), perplexity = p)
    plot(tsne_out$Y)
  }
}


```

in: ion data
    optimal perplexity
out: (x,y) coordinates
```{r echo = FALSE, message = FALSE}

run_tSNE <- function(data, optimal_p) {
  all_tsne_out <- Rtsne(as.matrix(data), perplexity = optimal_p)
  tsne_coords <- as.data.frame(all_tsne_out$Y)
  return(tsne_coords)
}


```


# PCA
in: ion data
out: (x,y) coordinates
```{r echo = FALSE, message = FALSE}

run_PCA <- function(data) {
  res <- prcomp(data)$x
  return(res[,c(1,2)])
}


``` 
     
     
# Create enrichment score summary for each method
in: list of pathways (names) of interest
    coordinates and clusters (results of cluster_data)
    metabolite data
    ions_inpath
out: summary of pathway, cluster, Q, Q_extreme, # sd away from mean
```{r echo = FALSE, method = FALSE}

es_summary <- function(pathways, clusters, inpath_matrix) {
  summary <- data.frame(matrix(ncol = 5, nrow = length(pathways)*length(unique(clusters$clusters))))
  names(summary) <- c("pathway", "cluster", "Q", "Q_extreme", "probability")
  i <- 1
  for (p in pathways) {
    for (c in unique(clusters$clusters)){
      df <- train[which(clusters$clusters == c),]
      res <- calc_ES(x_array = colMeans(df), pathway = p, ions_inpath = inpath_matrix, plot = TRUE, cluster_description = c)
      summary$pathway[i] <- p
      summary$cluster[i] <- c
      summary$Q[i] <- res$Q
      summary$Q_extreme[i] <- res$Q_extreme
      summary$probability[i] <- res$probability
      i <- i + 1
    }
  }
  
  return(summary)
  
}


```
























     
     
     
    