---
title: "Pre - Processing"
output: word_document
---

upload clinical data
```{r}
rawdata <- read.csv("../data/RedCap.csv")
```
subset initial visit only
```{r}
df_init <- rawdata[,grepl("v3y0", colnames(rawdata))]

```

filter patients based on inclusion criteria in subsetted data and whole data
```{r}
# drop patients who have had an amputation
df_init <- df_init[-grep(1, df_init$amputation_v3y0),]
rawdata <- rawdata[-grep(1, rawdata$amputation_v3y0),]


# drop patients who did not give genetic consent
df_init <- df_init[-grep(0, df_init$genetic_consent_bl_v3y0),]
rawdata <- rawdata[-grep(0, rawdata$genetic_consent_bl_v3y0),]

# drop patients using SERMS
df_init <- df_init[-grep(1, df_init$serms_v3y0),]
rawdata <- rawdata[-grep(1, rawdata$serms_v3y0),]

```

drop features that are not useful in initial data set and in raw data
inclusion criteria: no amputations, genetic consent given, not taking SERMS
```{r}
ignore <- c()
# drop features that contain > 20% NA or missing values in the initial data set and record which features they are

for (factor in 1:ncol(df_init)) {
  blanks <- sum(is.na(df_init[,factor])) # number of NA entries
  if (blanks == 0) blanks <-  nrow(df_init[df_init[,factor] == "",]) # number of blank entries
  total <- length(df_init[,factor]) # total number of entries
  if (blanks/total >= .2) { ignore <- c(ignore, factor) } # ignore factor if >= 20% of entries are blank
}

# drop calculated features
# fram
ignore <- c(ignore, grep("fram", names(df_init)))

# survival analysis
ignore <- c(ignore, grep("sa_", names(df_init)))
ignore <- c(ignore, grep("time_", names(df_init)))
ignore <- c(ignore, grep("scr_roche", names(df_init)))

# side by side time
ignore <- c(ignore, grep("side_by_side", names(df_init)))

# drop screening factors (should be the same across the board now)
ignore <- c(ignore, grep("amputation", names(df_init)))
ignore <- c(ignore, grep("genetic_consent", names(df_init)))
ignore <- c(ignore, grep("serms", names(df_init)))

# drop ions
ignore <- c(ignore, grep("ion_", names(df_init)))

# drop certain categorical variables based on distribution
ignore <- c(ignore, grep("albuminuria", names(df_init)))
ignore <- c(ignore, grep("diabetes", names(df_init)))
ignore <- c(ignore, grep("roche", names(df_init)))
ignore <- c(ignore, grep("hibp", names(df_init)))
ignore <- c(ignore, grep("igfr_baseline", names(df_init)))
ignore <- c(ignore, grep("egfr", names(df_init)))
ignore <- c(ignore, grep("esrd", names(df_init)))


df_init <- df_init[-ignore]


```

summarize remaining features
```{r}
library(knitr)

cont_df <- data.frame(matrix(nrow = nrow(df_init), ncol= 4, 0))
colnames(cont_df) <- c("name", "n", "mean", "std")
i <- 1
for (factor in 1:ncol(df_init)) {
  if (length(table(df_init[,factor])) > 10) {  # factor is continuous
    cont_df$name[i] <- colnames(df_init)[factor]
    cont_df$n[i] <- sum(!is.na(df_init[,factor]))
    cont_df$mean[i] <- mean(df_init[,factor], na.rm = TRUE)
    cont_df$std[i] <- sd(df_init[,factor], na.rm = TRUE)
    i <- i + 1
  } else { # factor is categorical
    cat_summary <- as.data.frame(table(df_init[,factor], useNA = "always"))
    cat_summary$Percentage <- (cat_summary$Freq*100)/total
    print(kable(cat_summary, caption = colnames(df_init)[factor]))
  }
}
cont_df <- cont_df[cont_df$name != 0,]
print(kable(cont_df, caption = "continuous variable summary"))

```

calculate EGFR rate and variance for each patient
```{r}
entry <- c(1:6)

#get number of unique visits on file
visits<- unique(substring(names(rawdata)[grep(paste("^egfr_cric", "_v", "[0-9]+", "y", "[0-9]+", sep = ""), names(rawdata))],11))

#init average yearly egfr exponential rate and volatility
Qbar <- rep(0,dim(rawdata)[1])
Qvar <- Qbar

#init valid visit counter
count <- Qbar

#get baseline egfr
Q0 <- rawdata$egfr_cric_v3y0
#get baseline time
t0 <- rawdata$days_thisvisit_v3y0

#loop over visits to calculate egfr return for every year
for (visit in visits[2:length(visits)]){
  #get current year egfr
  Qt <- rawdata[,paste("egfr_cric_",visit,sep = "")]
  #get time this visit
  t <- rawdata[,paste("days_thisvisit_",visit,sep = "")]

  #calculate egfr return in units of per year
  #Qdot <- 365*log(Qt/Q0)/(t-t0)
  #calculate egfr return in units of per visit
  Qdot <- log(Qt/Q0)

  #print(visit)
  #print(Qdot)
  
  #get the valid visit indecies
  vidx <- which(!is.na(Qdot))
  count[vidx] <- count[vidx] + 1 
  
  #accumulate first and second moments
  Qbar[vidx] <- Qbar[vidx]+Qdot[vidx]
  Qvar[vidx] <- Qvar[vidx]+Qdot[vidx]*Qdot[vidx]
    
  #update previous time and egfr
  Q0 <- Qt
  t0 <- t
}

#calculate mean and variance
Qbar <- Qbar/count
Qvar <- Qvar/count - Qbar*Qbar

#update rawdata
rawdata$Qbar <- Qbar
rawdata$Qvar <- Qvar

QQbar<- quantile(rawdata$Qbar,probs = seq(0,1,.25), na.rm = TRUE)
QQvar<- quantile(rawdata$Qvar,probs = seq(0,1,.25), na.rm = TRUE)

#init long form data for egfr with baseline visit
df <- rawdata[1:2]
df$rgroup <- findInterval(rawdata$Qbar, QQbar)
df$vgroup <- findInterval(rawdata$Qvar, QQvar)
df$mu_egfr_rate <- rawdata$Qbar
df$sigma_egfr_rate <-rawdata$Qvar
visit <- visits[1]
df$egfr <- rawdata[,paste("egfr_cric_",visit,sep = "")]
df$year <- 0
#loop over visits
for (i in c(2:length(visits))){
  df2 <- rawdata[1:2]
  df2$rgroup <- findInterval(rawdata$Qbar, QQbar)
  df2$vgroup <- findInterval(rawdata$Qvar, QQvar)
  df2$mu_egfr_rate <- rawdata$Qbar
  df2$sigma_egfr_rate <-rawdata$Qvar
  df2$egfr <- rawdata[,paste("egfr_cric_",visits[i],sep = "")]
  df2$year <- i
  df <- rbind(df,df2)
}




```

impute missing values 
```{r}
library(DMwR)

res <- knnImputation(df_init, k = 10, meth = "median", scale = F)

```
drop all features that we dropped from the first visit in all the other visits
```{r}
ig_raw <- c()

for (f in 1:length(names(df_init)[ignore])) {
  lookfor <- gsub("v3y0", "", names(df_init)[ignore][f])
  ig_raw <- c(ig_raw, grep(lookfor, names(rawdata)))
}
rawdata <- rawdata[-!is.na(ig_raw)]
```

normalize data with box-cox transformation

```{r}
library(car)

# if lambda needed to transform data is too high and there is not a significan change in normality, do not transform the data
summary <- data.frame("init" = c(1:200), "name" = 0, "initial_rsquared" = 0, "transformed_rsquared" = 0, "lambda" = 0, "transformed" = 1, "n_missing" = 0)
summary <- summary[-c(1)]
line <- 1

biglambda <- c()
untransformed <- c()
for (factor in 2:ncol(res)) {
  if (length(table(res[,factor])) > 10) {  # factor is continuous
    my_object <- res[,factor]
    name <- colnames(res)[factor]
        
    # before transformation

    qq_before <- qqnorm(my_object, ylab=name, main = paste0(name, " before transformation"), plot.it = FALSE)
    r_before <- cor(qq_before$x,qq_before$y)

    bc_result <- boxCox(my_object ~ 1, plotit = F, lambda = seq(-5,5,.1), eps = 0, family = "yjPower")
    bc_df <- data.frame(bc_result$x, bc_result$y)
    lambda <- bc_df[with(bc_df, order(-bc_df$bc_result.y)),][1,1]
        
    # after transformation
    if (lambda == 0) {transformed <- log(my_object)}
    if (lambda != 0) {transformed <- (my_object ^ lambda -1) / lambda}
          
    qq_after <- qqnorm(transformed, ylab=name, main = paste0(name, " after transformation"), plot.it = FALSE)
    r_after <- cor(qq_after$x,qq_after$y, use = "na.or.complete")

    if ((lambda > 1) && ((r_after - r_before)/r_before < .05)) {   # if lambda > 1 but there is no significant shift towards normal, use original data
      transformed <- my_object
      summary$transformed[line] <- 0
      untransformed <- c(untransformed, name)           # note exceptions
    } else {
      res[,factor] <- transformed 
      if (lambda> 1) {biglambda <- c(biglambda, name)} #note big lambdas
    }
        
    summary$name[line] <- name
    summary$initial_rsquared[line] <- r_before*r_before
    summary$transformed_rsquared[line] <- r_after*r_after
    summary$lambda[line] <- lambda
    summary$n_missing[line] <- nrow(res) - length(transformed)
    line <- line + 1
  }
}

# note these exceptions
print(untransformed)
```

output data
```{r}
write.csv(res, "../data/preprocessingresults.csv", row.names = FALSE)
write.csv(df, "../data/egfr_results.csv", row.names = FALSE)

```