---
title: "Pre - Processing"
output: word_document
---

# Objectives
* Filter data
  + Screen patients
  + Drop factors
* Calculate EGFR rate for each patient
* Impute missing values
* Normalize data
  + Box-Cox transformation for continuous variables
  + Convert categorical variables into binary


```{r echo = FALSE, results = 'asis'}
# upload raw data; clinical features located on columns 7 - 2271
rawdata <- read.csv("../data/RedCap.csv")
rawdata <- rawdata[c(2, 12:2271)] # keep patient ID and clinical features


# untargeted CRIC metadata
untarmet <- read.csv("../data/clinical_creatinine_normalized_data/CRIC_clinical_creatinine_normalized_samplemetadata.csv", header=FALSE, stringsAsFactors = FALSE)

```


```{r echo = FALSE, results = 'asis'}
is.continuous <- function(df, f){
  return(length(table(df[,f])) > 10)
}

factorize <- function(df){
  for (factor in 1:ncol(df)) {
    if (length(table(df[,factor])) == 2) {
      df[,factor] <- as.factor(df[,factor])
    }
  }
  return(df)
}


```

```{r echo = FALSE, results = 'asis'}
# extract patient IDs from untargeted data
vid <- substr(untarmet$V2,1,7)
pid <- unique(vid)

# remove patients in rawdata who are not represented in untarmet, and vice versa
if (any(!pid %in% rawdata$patientid)) {print("there are patients in the metadata that are not in the raw data")}

no_match <- row.names(rawdata[!rawdata$patientid %in% pid,])
rawdata <- rawdata[-as.integer(no_match),]
if (any(!rawdata$patientid %in% pid)) {print("there are patients in the raw data that are not in the metadata")}   #test

#subset initial visit only
df_init <- data.frame(rawdata$patientid) # include patient id information
df_init <- data.frame(df_init, rawdata[,grepl("v3y0", colnames(rawdata))]) # find all factors from v3y0
```

# Filter patients based on inclusion criteria
```{r echo = FALSE, results = 'asis'}
drop_patients <- c()

# amputation
amputee_ids <- df_init[grep(1, df_init$amputation_v3y0),]$rawdata.patientid

# genetic consent
no_consent_ids <- df_init[grep(0, df_init$genetic_consent_bl_v3y0),]$rawdata.patientid

# SERMS
SERMS_ids <- df_init[grep(1, df_init$serms_v3y0),]$rawdata.patientid

```

EGFR rate informs inclusion criteria; only include patients who have 4+ EGFR readings
```{r echo = FALSE, results = 'asis'}

# Calculate EGFR rate and variance for each patient
entry <- c(1:6)

#get number of unique visits on file
visits<- unique(substring(names(rawdata)[grep(paste("^egfr_cric", "_v", "[0-9]+", "y", "[0-9]+", sep = ""), names(rawdata))],11))

#init average yearly egfr exponential rate and volatility
Qbar <- rep(0,dim(rawdata)[1])
Qvar <- Qbar

#init valid visit counter
count <- Qbar

#get baseline egfr
Q0 <- rawdata$egfr_cric_v3y0
#get baseline time
t0 <- rawdata$days_thisvisit_v3y0

#loop over visits to calculate egfr return for every year
for (visit in visits[2:length(visits)]){
  #get current year egfr
  Qt <- rawdata[,paste("egfr_cric_",visit,sep = "")]
  #get time this visit
  t <- rawdata[,paste("days_thisvisit_",visit,sep = "")]

  #calculate egfr return in units of per year
  #Qdot <- 365*log(Qt/Q0)/(t-t0)
  #calculate egfr return in units of per visit
  Qdot <- log(Qt/Q0)

  #print(visit)
  #print(Qdot)
  
  #get the valid visit indecies
  vidx <- which(!is.na(Qdot))
  count[vidx] <- count[vidx] + 1 
  
  #accumulate first and second moments
  Qbar[vidx] <- Qbar[vidx]+Qdot[vidx]
  Qvar[vidx] <- Qvar[vidx]+Qdot[vidx]*Qdot[vidx]
    
  #update previous time and egfr
  Q0 <- Qt
  t0 <- t
}

#calculate mean and variance
Qbar <- Qbar/count
Qvar <- Qvar/count - Qbar*Qbar

#update rawdata
rawdata$Qbar <- Qbar
rawdata$Qvar <- Qvar

QQbar<- quantile(rawdata$Qbar,probs = seq(0,1,.25), na.rm = TRUE)
QQvar<- quantile(rawdata$Qvar,probs = seq(0,1,.25), na.rm = TRUE)


#init long form data for egfr with baseline visit
df <- rawdata[1:2]
df$rgroup <- findInterval(rawdata$Qbar, QQbar)
df$vgroup <- findInterval(rawdata$Qvar, QQvar)
df$mu_egfr_rate <- rawdata$Qbar
df$sigma_egfr_rate <-rawdata$Qvar
visit <- visits[1]
df$egfr <- rawdata[,paste("egfr_cric_",visit,sep = "")]
df$year <- 0
#loop over visits
for (i in c(2:length(visits))){
  df2 <- rawdata[1:2]
  df2$rgroup <- findInterval(rawdata$Qbar, QQbar)
  df2$vgroup <- findInterval(rawdata$Qvar, QQvar)
  df2$mu_egfr_rate <- rawdata$Qbar
  df2$sigma_egfr_rate <-rawdata$Qvar
  df2$egfr <- rawdata[,paste("egfr_cric_",visits[i],sep = "")]
  df2$year <- i
  df <- rbind(df,df2)
}

# delete df2 
rm(df2)

# add egfr information to raw data
egfr_v0 <- df[df$year == 0,]
rawdata$mu_egfr <- egfr_v0$mu_egfr_rate
rawdata$sigma_egfr <- egfr_v0$sigma_egfr_rate
rawdata$r_group <- egfr_v0$rgroup
rawdata$v_group <- egfr_v0$vgroup


# drop patients with less than 4 egfr readings
egfr_readings <- count + 1
rawdata$egfr_readings <- egfr_readings
few_egfr_ids <- rawdata[rawdata$egfr_readings < 4,]$patientid


```

```{r echo = FALSE, results = 'asis'}

# drop patients with > 20% missing from initial visit
missing20_ids <- c()
for (p in 1:nrow(df_init)) {
  if (sum(is.na(df_init[p,])) / ncol(df_init) > .2) {
    missing20_ids <- c(missing20_ids, df_init(p,1))
  }
}


# drop the patients from the raw data set
rawdata <- rawdata[!rawdata$patientid %in% amputee_ids,]
rawdata <- rawdata[!rawdata$patientid %in% few_egfr_ids,]
rawdata <- rawdata[!rawdata$patientid %in% missing20_ids,]
rawdata <- rawdata[!rawdata$patientid %in% no_consent_ids,]
rawdata <- rawdata[!rawdata$patientid %in% SERMS_ids,]


# update the initial year
df_init <- data.frame(rawdata$patientid) # include patient id information
df_init <- data.frame(df_init, rawdata[,grepl("v3y0", colnames(rawdata))]) # find all factors from v3y0


# save egfr data
egfr_data <- data.frame(rawdata$mu_egfr, rawdata$sigma_egfr, rawdata$r_group, rawdata$v_group)

# save patient data
patients <- rawdata$patientid

```


Drop features that have been calculated such as survival analyses, were used for screening, are missing more than 20% of entries, etc.
```{r echo = FALSE, results = 'asis'}

#drop features that are not useful in initial data set and in raw data
ignore <- c()

# drop features that contain > 20% NA or missing values in the initial data set and record which features they are
for (factor in 1:ncol(df_init)) {
  blanks <- sum(is.na(df_init[,factor])) # number of NA entries
  if (blanks == 0) blanks <-  nrow(df_init[df_init[,factor] == "",]) # number of blank entries
  total <- length(df_init[,factor]) # total number of entries
  if (blanks/total >= .2) { ignore <- c(ignore, factor) } # ignore factor if >= 20% of entries are blank
}

# drop calculated features
# fram
ignore <- c(ignore, grep("fram", names(df_init)))

# survival analysis
ignore <- c(ignore, grep("sa_", names(df_init)))
ignore <- c(ignore, grep("time_", names(df_init)))
ignore <- c(ignore, grep("scr_roche", names(df_init)))

# side by side time
ignore <- c(ignore, grep("side_by_side", names(df_init)))

# drop screening factors (should be the same across the board now)
ignore <- c(ignore, grep("amputation", names(df_init)))
ignore <- c(ignore, grep("genetic_consent", names(df_init)))
ignore <- c(ignore, grep("serms", names(df_init)))

# drop ions (if present)
ignore <- c(ignore, grep("ion_", names(df_init)))

# drop certain categorical variables based on distribution
ignore <- c(ignore, grep("albuminuria", names(df_init)))
ignore <- c(ignore, grep("diabetes", names(df_init)))
ignore <- c(ignore, grep("roche", names(df_init)))
ignore <- c(ignore, grep("hibp", names(df_init)))
ignore <- c(ignore, grep("igfr_baseline", names(df_init)))
ignore <- c(ignore, grep("egfr", names(df_init)))
ignore <- c(ignore, grep("esrd", names(df_init)))

# drop u 24h proteins because they are measured in urinary_protein
ignore <- c(ignore, grep("ualbumin", names(df_init)))
ignore <- c(ignore, grep("ucreatinine", names(df_init)))
ignore <- c(ignore, grep("uprotein", names(df_init)))



#drop all features that we dropped from the first visit in all the other visits
ig_raw <- c()

for (f in 1:length(names(df_init)[ignore])) {
  lookfor <- gsub("v3y0", "", names(df_init)[ignore][f])
  ig_raw <- c(ig_raw, grep(lookfor, names(rawdata)))
}

ig_raw <- unique(ig_raw)
rawdata <- rawdata[-ig_raw]


# drop ignore features from df_init
df_init <- df_init[-ignore]
print(paste0("Drop ", length(ignore), " features."))


```
Convert categorical data to binary
```{r echo = FALSE, message = FALSE, results = 'asis'}

#Turn all categorical variables into binary variables
cat_to_binary <- function(df) {
  n_factors <- length(df)
  colnames <- colnames(df)
  for (factor in 2:ncol(df)) {                                                       # for each factor
    if ((length(table(df[,factor])) < 10) && (length(table(df[,factor])) > 2)) {    # if factor is categorical and not already binary
      new_names <- levels(as.data.frame(table(df[,factor]))$Var1)                    # save category names as new_names
      for (name in 1:length(new_names)) {                                             # for each of these category names
        df[,n_factors + 1] <- as.integer(grepl(name,df[,factor]))                   # add a column to df with binary form of the cat
        colnames <- c(colnames, paste0(colnames[factor], "_", name))                  # update df column names
        n_factors <- n_factors + 1                                                    # update number of factors in df
      }
      df <- df[-factor]                                                             # remove original categorical factor
      colnames <- colnames[-factor]                                                   # update colnames
      n_factors <- n_factors - 1                                                      # update number of factors in df
    }
  }
  colnames(df) <- colnames
  return(df)
}

df_init <- cat_to_binary(df_init)
rawdata <- cat_to_binary(rawdata)

```

Random Forest as a validation for factor dropping later. exclude nas
```{r echo = FALSE, results = 'asis', message = FALSE}
library(randomForest)
library(knitr)
df_rf <- factorize(df_init)
df_rf <- df_rf[-1]
df_rf <- data.frame(df_rf, egfr_data$rawdata.mu_egfr)

rf.test <- randomForest(df_rf$egfr_data.rawdata.mu_egfr ~ . , data = df_rf, importance = TRUE, na.action = "na.exclude")
kable(importance(rf.test, type = 1))

```





Drop features that do not predict mu EGFR well 
+ linear model for continuous factors
+ wilcox for categorical
```{r echo = FALSE, results = 'asis', message = FALSE}
library(knitr)
# continuous vars
df_init <- data.frame(df_init, egfr_data$rawdata.mu_egfr)


df_cont <- data.frame("init" = c(1:ncol(df_init)), "name" = 0, 
                        "B" = 0, "p" = 0, "r_squared" = 0)

not_associated <- c()

line <- 1
for (factor in 2:ncol(df_init)) {
  if (is.continuous(df_init, factor)) {  
    df_cont$name[line] <- colnames(df_init)[factor]
      
    lm.res <- lm(df_init$egfr_data.rawdata.mu_egfr ~ df_init[,factor])
      
    df_cont$B[line] <- lm.res$coefficients[2]
    df_cont$p[line] <- summary(lm.res)$coefficients[,4][2] 
    df_cont$r_squared[line] <- summary(lm.res)$r.squared
    
    if (df_cont$p[line] >= 0.05) {not_associated <- c(not_associated, factor)}
    
    line <- line + 1
    
  }
}
  
df_cont <- df_cont[df_cont$name != 0,]
df_cont <- df_cont[-1]
  
print(kable(df_cont), label = description)

cont_drop <- length(not_associated)

# categorical vars
df_cat <- data.frame("init" = c(1:ncol(df_init)), "name" = 0, "pval" = 0, "n" = 0, "percent" = 0)
df_cat <- df_cat[-1]
line <- 1

for (factor in 2:ncol(df_init)) {
  if ((length(table(df_init[,factor])) < 10) && (length(table(df_init[,factor])) > 1)) {  #
    df_cat$name[line] <- colnames(df_init)[factor]  
    
    one <- df_init[df_init[,factor] == 1,]$egfr_data.rawdata.mu_egfr
    not_one <- df_init[df_init[,factor] != 1,]$egfr_data.rawdata.mu_egfr
    
    df_cat$pval[line] <- wilcox.test(one, not_one)$p.value
    cat_summary <- as.data.frame(table(df_init[,factor], useNA = "always"))
    df_cat$n[line] <- cat_summary[cat_summary$Var1==1,]$Freq[1]
    df_cat$percent[line] <- df_cat$n[line]/sum(cat_summary$Freq) * 100
      
    if(!grepl("sex", names(df_init)[factor])) {if (df_cat$pval[line] >= 0.05) {not_associated <- c(not_associated, factor)}}  # save sex even if not significant
  
    
    line <- line + 1
    
  }
}
    
df_cat <- df_cat[df_cat$name != 0,]

kable(df_cat)

cat_drop <- length(not_associated)-cont_drop

# drop some variiables due to redundancy after looking at codes in cric codebook

# drop race cat 1 and race ethnicity because they are  redundant nad does not contain as much info as race_ethnicity_cat2
not_associated<- c(not_associated, grep("race_cat_1", names(df_init)))
not_associated<- c(not_associated, grep("race_ethnicity_v3y0", names(df_init)))
# drop edu cat 2 and edu cat 1 bc redundant
not_associated<- c(not_associated, grep("edu_cat_2", names(df_init)))
not_associated<- c(not_associated, grep("edu_cat_1", names(df_init)))

# dropping features that are not related to mu egfr

#drop all features that we dropp from the first visit in all the other visits
not_assoc_raw <- c()

for (f in 1:length(names(df_init)[not_associated])) {
  lookfor <- gsub("v3y0", "", names(df_init)[ignore][f])
  not_assoc_raw <- c(not_assoc_raw, grep(lookfor, names(rawdata)))
}

not_assoc_raw <- !is.na(unique(not_assoc_raw))
rawdata <- rawdata[-not_assoc_raw]

df_init <- df_init[-not_associated]
print(paste0(cont_drop, " continuous features were dropped because they did not predict egfr well enough."))
print(paste0(cat_drop, " categorical features were dropped because they did not predict egfr well enough."))


```

# Summary of remaining features:
```{r echo = FALSE, results = 'asis'}
library(knitr)

cont_df <- data.frame(matrix(nrow = 100, ncol= 4, 0))
colnames(cont_df) <- c("name", "n", "mean", "std")
i <- 1
for (factor in 2:ncol(df_init)) {
  if (length(table(df_init[,factor])) > 10) {  # factor is continuous
    cont_df$name[i] <- colnames(df_init)[factor]
    cont_df$n[i] <- sum(!is.na(df_init[,factor]))
    cont_df$mean[i] <- mean(df_init[,factor], na.rm = TRUE)
    cont_df$std[i] <- sd(df_init[,factor], na.rm = TRUE)
    i <- i + 1
  } else { # factor is categorical
    cat_summary <- as.data.frame(table(df_init[,factor], useNA = "always"))
    cat_summary$Percentage <- (cat_summary$Freq*100)/total
    print(kable(cat_summary, caption = colnames(df_init)[factor]))
  }
}
cont_df <- cont_df[cont_df$name != 0,]
print(kable(cont_df, caption = "continuous variable summary"))

```





```{r echo = FALSE, results = 'asis', message=FALSE}
# Impute missing values using k-nearest neighbors
library(DMwR)
df_init <- df_init[-ncol(df_init)] # remove egfr data
res <- knnImputation(df_init[-1], k = 10, meth = "median", scale = F) # do not use patient id in imputation. use median because there are no NAs in categorical vars, but would want weighavg for those

# save a copy of non imputed data
non_imputed <- df_init[-1]

```



```{r echo = FALSE, results = 'asis'}

# add egfr information to non imputed data set
non_imputed$mu_egfr <- egfr_data$rawdata.mu_egfr
non_imputed$sigma_egfr <- egfr_data$rawdata.sigma_egfr
non_imputed$r_group <- egfr_data$rawdata.r_group
non_imputed$v_group <- egfr_data$rawdata.v_group

# remove all rows with NA
non_imputed <- non_imputed[complete.cases(non_imputed),]

# extract non_imputed_egfr data and remove for transformation
non_imputed_egfr_data <- non_imputed[30:33]
non_imputed <- non_imputed[-c(30:33)]

```


```{r echo = FALSE, results = 'asis', message = FALSE}
#Normalize continuous data with box-cox transformation
transform_continuous_data <- function(df, description){

  library(car)
  
  # if lambda needed to transform data is too high and there is not a significant change in normality, do not transform the data
  summary <- data.frame("init" = c(1:200), "name" = 0, "initial_rsquared" = 0, "transformed_rsquared" = 0, "lambda" = 0, "transformed" = 1, "n_missing" = 0)
  summary <- summary[-c(1)]
  line <- 1
  
  biglambda <- c()
  untransformed <- c()
  for (factor in 2:ncol(df)) {
    if (length(table(df[,factor])) > 10) {  # factor is continuous
      my_object <- df[,factor]
      name <- colnames(df)[factor]
          
      # before transformation
  
      qq_before <- qqnorm(my_object, ylab=name, main = paste0(name, " before transformation"), plot.it = FALSE)
      r_before <- cor(qq_before$x,qq_before$y)
  
      bc_dfult <- boxCox(my_object ~ 1, plotit = F, lambda = seq(-5,5,.1), eps = 0, family = "yjPower")
      bc_df <- data.frame(bc_dfult$x, bc_dfult$y)
      lambda <- bc_df[with(bc_df, order(-bc_df$bc_dfult.y)),][1,1]
          
      # after transformation
      if (lambda == 0) {transformed <- log(my_object)}
      if (lambda != 0) {transformed <- (my_object ^ lambda -1) / lambda}
            
      qq_after <- qqnorm(transformed, ylab=name, main = paste0(name, " after transformation"), plot.it = FALSE)
      r_after <- cor(qq_after$x,qq_after$y, use = "na.or.complete")
  
      if ((lambda > 1) && ((r_after - r_before)/r_before < .05)) {   # if lambda > 1 but there is no significant shift towards normal, use original data
        transformed <- my_object
        summary$transformed[line] <- 0
        untransformed <- c(untransformed, name)           # note exceptions
      } else {
        df[,factor] <- transformed 
        if (lambda> 1) {biglambda <- c(biglambda, name)} #note big lambdas
      }
          
      summary$name[line] <- name
      summary$initial_rsquared[line] <- r_before*r_before
      summary$transformed_rsquared[line] <- r_after*r_after
      summary$lambda[line] <- lambda
      summary$n_imputed[line] <- sum(is.na(df_init[,factor]))
      line <- line + 1
    }
  }
  
  summary <- summary[summary$name != 0,]
  
  # note these exceptions
  print(kable(summary, label = description))
  print("The following continuous factors were not transformed because transformation did not significantly shift them towards normality: ")
  print(untransformed)
  
  return(df)
}
```


```{r echo = FALSE, results = "asis"}
# transform the data

res <- transform_continuous_data(res, "imputed data")
non_imputed <- transform_continuous_data(non_imputed, "non imputed data")

```




```{r echo = FALSE, results = 'asis'}
#add egfr values back to result and to non_imputed data 
res <- data.frame(res, egfr_data)

non_imputed <- data.frame(non_imputed, non_imputed_egfr_data)

```



```{r}
test.res <- lm(mu_egfr ~ urine_albumin_v3y0 + urine_creatinine_v3y0, data = non_imputed )
print(summary(test.res)$r.squared)
```



  
# Compare Imputed and Non Imputed Data Sets
use a linear model to predict egfr rate
```{r, echo = FALSE, message = FALSE, results = 'asis'}

imputed_lm.res <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group, data = res)
print(paste0("imputed data r squared: ", summary(imputed_lm.res)$r.squared))

nonimputed_lm.res <- lm(mu_egfr ~ . -sigma_egfr - v_group - r_group, data = non_imputed)
print(paste0("non imputed data r squared: ", summary(nonimputed_lm.res)$r.squared))

print(paste0("using non imputed data requires ", nrow(res) - nrow(non_imputed), " patients to be dropped due to missing values."))

```

```{r results='asis', echo = FALSE, message = FALSE}
# based on clinical features, make list of all patient IDs that were included
# use included patients to filter untargeted data

# load untargeted cric data
#untar <- read.csv("../data/clinical_creatinine_normalized_data/CRIC_clinical_creatinine_normalized_data.csv", header=FALSE)

#t_untar <- as.data.frame(t(untar))

#t_untar$vid <- vid
#t_untar <- t_untar[t_untar$vid %in% patients,]
#t_untar <- t_untar[-15435]


# save untargeted data filtered by patient
#filtered_untar <- as.data.frame(t(t_untar))
#write.csv(filtered_untar, "../data/clinical_creatinine_normalized_data/cric_normalized_data_filteredByPatient.csv", row.names = FALSE)

```


```{r echo = FALSE, results = 'asis'}
#save data
write.csv(res, "../data/preprocessingresults.csv", row.names = FALSE)
write.csv(non_imputed, "../data/nonimputed_preprocessingresults.csv", row.names = FALSE)

```