---
title: "Pre - Processing"
output: word_document
---

# Objectives
* Filter data
  + Screen Patients
  + Feature Selection 1
  + Feature Selection 2
* Calculate Outcomes: EGFR rate and Volatility
* Impute missing values
* Normalize data
  + Continuous Variables: Box-Cox transformation and unit scale
  + Categorical Variables: Convert to binary


```{r echo = FALSE, results = 'asis'}
# upload raw data; clinical features located on columns 7 - 2271
rawdata <- read.csv("../data/RedCap.csv")
rawdata <- rawdata[c(2, 12:2271)] # keep patient ID and clinical features


# untargeted CRIC metadata
untarmet <- read.csv("../data/CRIC_clinical_creatinine_normalized_samplemetadata.csv", header=FALSE, stringsAsFactors = FALSE)

```


```{r echo = FALSE, results = 'asis'}
# functions for later use
is.continuous <- function(df, f){
  return(length(table(df[,f])) > 10)
}

factorize <- function(df){
  for (factor in 1:ncol(df)) {
    if (length(table(df[,factor])) == 2) {
      df[,factor] <- as.factor(df[,factor])
    }
  }
  return(df)
}

cat_to_binary <- function(df) {
  df$sex_v3y0 <- df$sex_v3y0-1 # make sex binary
  n_factors <- length(df)
  colnames <- colnames(df)
  for (factor in 2:ncol(df)) {                                                       # for each factor
    if ((length(table(df[,factor])) < 10) && (length(table(df[,factor])) > 2)) {    # if factor is categorical and not already binary
      new_names <- levels(as.data.frame(table(df[,factor]))$Var1)                    # save category names as new_names
      for (name in 1:length(new_names)) {                                             # for each of these category names
        df[,n_factors + 1] <- as.integer(grepl(name,df[,factor]))                   # add a column to df with binary form of the cat
        colnames <- c(colnames, paste0(colnames[factor], "_", name))                  # update df column names
        n_factors <- n_factors + 1                                                    # update number of factors in df
      }
      df <- df[-factor]                                                             # remove original categorical factor
      colnames <- colnames[-factor]                                                   # update colnames
      n_factors <- n_factors - 1                                                      # update number of factors in df
    }
  }
  colnames(df) <- colnames
  return(df)
}

```

```{r echo = FALSE, results = 'asis'}
# extract patient IDs from untargeted data
vid <- substr(untarmet$V2,1,7)
pid <- unique(vid)

# remove patients in rawdata who are not represented in untarmet, and vice versa
if (any(!pid %in% rawdata$patientid)) {print("there are patients in the metadata that are not in the raw data")}

no_match <- row.names(rawdata[!rawdata$patientid %in% pid,])
rawdata <- rawdata[-as.integer(no_match),]
if (any(!rawdata$patientid %in% pid)) {print("there are patients in the raw data that are not in the metadata")}   #test

```

# Filter patients based on inclusion criteria
```{r echo = FALSE, results = 'asis'}
# amputation
rawdata <- rawdata[rawdata$amputation_v3y0 == 0,]

# genetic consent
rawdata <- rawdata[rawdata$genetic_consent_bl_v3y0 == 1,]

# SERMS
rawdata <- rawdata[rawdata$serms_v3y0 == 0,]

```

EGFR rate informs inclusion criteria; only include patients who have 4+ EGFR readings
```{r echo = FALSE, results = 'asis'}

# Calculate EGFR rate and variance for each patient
entry <- c(1:6)

#get number of unique visits on file
visits<- unique(substring(names(rawdata)[grep(paste("^egfr_ckd_epi", sep = ""), names(rawdata))],14))

#init average yearly egfr exponential rate and volatility
Qbar <- rep(0,dim(rawdata)[1])
Qvar <- Qbar

#init valid visit counter
count <- Qbar

#get baseline egfr
Q0 <- rawdata$egfr_ckd_epi_v3y0

#get baseline time
t0 <- rawdata$days_thisvisit_v3y0

#loop over visits to calculate egfr return for every year
for (visit in visits[2:length(visits)]){
  Qt <- NA
  t <- NA
  
  #get current year egfr
  Qt <- rawdata[,paste("egfr_ckd_epi_",visit,sep = "")]
  #get time this visit
  t <- rawdata[,paste("days_thisvisit_",visit,sep = "")]

  #calculate egfr return in units of per year
  #Qdot <- 365*log(Qt/Q0)/(t-t0)
  #calculate egfr return in units of per visit
  Qdot <- log(Qt/Q0)

  #print(visit)
  #print(Qdot)
  
  #get the valid visit indecies
  vidx <- which(!is.na(Qdot))
  count[vidx] <- count[vidx] + 1 
  
  #accumulate first and second moments
  Qbar[vidx] <- Qbar[vidx]+Qdot[vidx]
  Qvar[vidx] <- Qvar[vidx]+Qdot[vidx]*Qdot[vidx]
    
  #update previous time and egfr
  Q0 <- Qt
  t0 <- t
}

#calculate mean and variance
Qbar <- Qbar/count
Qvar <- Qvar/count - Qbar*Qbar

#update rawdata
rawdata$Qbar <- Qbar
rawdata$Qvar <- Qvar

QQbar<- quantile(rawdata$Qbar,probs = seq(0,1,.25), na.rm = TRUE)
QQvar<- quantile(rawdata$Qvar,probs = seq(0,1,.25), na.rm = TRUE)
```


```{r}

#init long form data for egfr with baseline visit
df <- rawdata[1:2]
df$rgroup <- findInterval(rawdata$Qbar, QQbar)
df$vgroup <- findInterval(rawdata$Qvar, QQvar)
df$mu_egfr_rate <- rawdata$Qbar
df$sigma_egfr_rate <-rawdata$Qvar
visit <- visits[1]
df$egfr <- rawdata[,paste("egfr_ckd_epi_",visit,sep = "")]
df$year <- 0
#loop over visits
for (i in c(2:length(visits))){
  df2 <- rawdata[1:2]
  df2$rgroup <- findInterval(rawdata$Qbar, QQbar)
  df2$vgroup <- findInterval(rawdata$Qvar, QQvar)
  df2$mu_egfr_rate <- rawdata$Qbar
  df2$sigma_egfr_rate <-rawdata$Qvar
  df2$egfr <- rawdata[,paste("egfr_ckd_epi_",visits[i],sep = "")]
  df2$year <- i
  df <- rbind(df,df2)
}

# delete df2 
rm(df2)

# add egfr information to raw data
egfr_v0 <- df[df$year == 0,]
rawdata$mu_egfr <- egfr_v0$mu_egfr_rate
rawdata$sigma_egfr <- egfr_v0$sigma_egfr_rate
rawdata$r_group <- egfr_v0$rgroup
rawdata$v_group <- egfr_v0$vgroup


# drop patients with less than 3 intervals
egfr_readings <- count 
rawdata$egfr_readings <- egfr_readings
few_egfr_ids <- rawdata[rawdata$egfr_readings < 3,]$patientid


```


```{r echo = FALSE, results = 'asis'}

# save the initial year
df_init <- rawdata[,grepl("v3y0", colnames(rawdata))] # find all factors from v3y0
df_init$patientid <- rawdata$patientid # include patient id information
df_init <- df_init[c(ncol(df_init), 1:(ncol(df_init)-1))]

# drop patients with > 20% missing from initial visit
missing20_ids <- c()
for (p in 1:nrow(df_init)) {
  if (sum(is.na(df_init[p,])) / ncol(df_init) > .2) {
    missing20_ids <- c(missing20_ids, df_init[p,1])
  }
}

# drop the patients
df_init <- df_init[!df_init$patientid %in% few_egfr_ids,]
df_init <- df_init[!df_init$patientid %in% missing20_ids,]

# save egfr data
egfr_data <- data.frame(rawdata$patientid, rawdata$mu_egfr, rawdata$sigma_egfr, rawdata$r_group, rawdata$v_group)

egfr_data <- egfr_data[egfr_data$rawdata.patientid %in% df_init$patientid,]

```


Drop features that have been calculated such as survival analyses, were used for screening, are missing more than 20% of entries, etc.
```{r echo = FALSE, results = 'asis'}

library(knitr)

#drop features that are not useful 
ignore <- c()

# drop features that contain > 20% NA or missing values in the initial data set and record which features they are
for (factor in 1:ncol(df_init)) {
  blanks <- sum(is.na(df_init[,factor])) # number of NA entries
  if (blanks == 0) blanks <-  nrow(df_init[df_init[,factor] == "",]) # number of blank entries
  total <- length(df_init[,factor]) # total number of entries
  if (blanks/total >= .2) { ignore <- c(ignore, factor) } # ignore factor if >= 20% of entries are blank
}

# drop calculated features
# fram
ignore <- c(ignore, grep("fram", names(df_init)))

# survival analysis
ignore <- c(ignore, grep("sa_", names(df_init)))
ignore <- c(ignore, grep("time_", names(df_init)))
ignore <- c(ignore, grep("scr_roche", names(df_init)))

# side by side time
ignore <- c(ignore, grep("side_by_side", names(df_init)))

# drop screening factors (should be the same across the board now)
ignore <- c(ignore, grep("amputation", names(df_init)))
ignore <- c(ignore, grep("genetic_consent", names(df_init)))
ignore <- c(ignore, grep("serms", names(df_init)))

# drop ions (if present)
ignore <- c(ignore, grep("ion_", names(df_init)))

# drop certain categorical variables based on distribution
ignore <- c(ignore, grep("albuminuria", names(df_init))) # derived
ignore <- c(ignore, grep("diabetes", names(df_init))) # everyone has it
ignore <- c(ignore, grep("roche", names(df_init))) # derived
ignore <- c(ignore, grep("hibp", names(df_init))) # repetitive
ignore <- c(ignore, grep("igfr_baseline", names(df_init))) # we use egfr
#ignore <- c(ignore, grep("egfr", names(df_init))) # keep this because we associate vars with it at baseline
ignore<- c(ignore, grep("egfr_cric", names(df_init))) # we are using egfr ckd bc clinical standard
ignore <- c(ignore, grep("esrd", names(df_init)))

# drop u 24h proteins because they are measured in urinary_protein (may not be as reliable as extretion rate, which is grams/24hr)
ignore <- c(ignore, grep("ualbumin", names(df_init)))
ignore <- c(ignore, grep("ucreatinine", names(df_init)))
ignore <- c(ignore, grep("uprotein", names(df_init)))

# remove waist because it is highly correlated with BMI; fit to a linear model to drive point home
lm.res <- lm(waist_v3y0 ~ bmi_v3y0, df_init)
kable(data.frame("B" = lm.res$coefficients[2], "p" = summary(lm.res)$coefficients[,4][2] , "r squared" = summary(lm.res)$r.squared), label = "linear model: BMI as a function of waist")
ignore <- c(ignore, grep("waist", names(df_init)))

# remove MAP because it is calculated
ignore <- c(ignore, grep("map", names(df_init)))

# drop some education and race categorical vars due to redundancy
ignore<- c(ignore, grep("edu_cat_2", names(df_init)))
ignore<- c(ignore, grep("edu_cat_1", names(df_init)))
ignore<- c(ignore, grep("race_cat_1", names(df_init)))
ignore<- c(ignore, grep("race_ethnicity_v3y0", names(df_init)))

# drop ignore features from df_init
df_init <- df_init[-ignore]
print(paste0("Drop ", length(ignore), " features."))


```


```{r echo = FALSE, message = FALSE, results = 'asis'}
#Turn all categorical variables into binary variables
df_init <- cat_to_binary(df_init)
rawdata <- cat_to_binary(rawdata)

```


Drop features that do not predict baseline EGFR well 
+ linear model for continuous factors
+ wilcox for categorical
```{r echo = FALSE, results = 'asis', message = FALSE}
library(knitr)
# continuous vars

df_cont <- data.frame("init" = c(1:ncol(df_init)), "name" = 0, 
                        "B" = 0, "p" = 0, "r_squared" = 0)

signif <- c()
cont_drop <- c()

line <- 1
for (factor in 2:ncol(df_init)) {
  if (is.continuous(df_init, factor)) {  
    df_cont$name[line] <- colnames(df_init)[factor]
    lm.res <- lm(df_init$egfr_ckd_epi_v3y0 ~ df_init[,factor])
      
    df_cont$B[line] <- lm.res$coefficients[2]
    df_cont$p[line] <- summary(lm.res)$coefficients[,4][2] 
    df_cont$r_squared[line] <- summary(lm.res)$r.squared
    
    if (df_cont$p[line] >= 0.05) {
      signif <- c(signif, factor)
      cont_drop <- c(cont_drop, factor)
    }        
    line <- line + 1
  }
}
  
df_cont <- df_cont[df_cont$name != 0,]
df_cont <- df_cont[-1]
  
print(kable(df_cont), label = description)

# categorical vars
df_cat <- data.frame("init" = c(1:(ncol(df_init)-1)), "name" = 0, "pval" = 0, "n" = 0, "percent" = 0)
df_cat <- df_cat[-1]
line <- 1

cat_drop <- c()

for (factor in 2:ncol(df_init)) {
  if (length(table(df_init[,factor])) == 1) {} # drop factors wth no variation
  else if (!is.continuous(df = df_init, f = factor)) {  
    df_cat$name[line] <- colnames(df_init)[factor]  
    
    one <- df_init[df_init[,factor] == 1,]$egfr_ckd_epi_v3y0
    not_one <- df_init[df_init[,factor] != 1,]$egfr_ckd_epi_v3y0
    
    df_cat$pval[line] <- wilcox.test(one, not_one)$p.value
    cat_summary <- as.data.frame(table(df_init[,factor], useNA = "always"))
    df_cat$n[line] <- cat_summary[cat_summary$Var1==1,]$Freq[1]
    df_cat$percent[line] <- df_cat$n[line]/sum(cat_summary$Freq) * 100
      
    if (df_cat$pval[line] < 0.05) {
      signif <- c(signif, factor)
      cat_drop <- c(cat_drop, factor)
    }
    line <- line + 1
  }
}
    
df_cat <- df_cat[df_cat$name != 0,]

kable(df_cat)

print(paste0(length(cont_drop), " continuous features had a p value of less than 0.05 predicting baseline egfr with the linear model."))
print(paste0(length(cat_drop), " categorical features had a p value of less than 0.05 predicting baseline egfr with the wilcox test."))


```


Drop features that do not predict baseline A1C well 
+ linear model for continuous factors
+ wilcox for categorical
```{r echo = FALSE, results = 'asis', message = FALSE}
library(knitr)
# continuous vars

df_cont <- data.frame("init" = c(1:ncol(df_init)), "name" = 0, 
                        "B" = 0, "p" = 0, "r_squared" = 0)

cont_drop <- c()

line <- 1
for (factor in 2:ncol(df_init)) {
  if (is.continuous(df_init, factor)) {  
    df_cont$name[line] <- colnames(df_init)[factor]
    lm.res <- lm(df_init$hemoglobin_a1c_v3y0 ~ df_init[,factor])
      
    df_cont$B[line] <- lm.res$coefficients[2]
    df_cont$p[line] <- summary(lm.res)$coefficients[,4][2] 
    df_cont$r_squared[line] <- summary(lm.res)$r.squared
    
    if (df_cont$p[line] >= 0.05) {
      signif <- c(signif, factor)
      cont_drop <- c(cont_drop, factor)
    }        
    line <- line + 1
  }
}
  
df_cont <- df_cont[df_cont$name != 0,]
df_cont <- df_cont[-1]
  
print(kable(df_cont), label = description)

# categorical vars
df_cat <- data.frame("init" = c(1:(ncol(df_init)-1)), "name" = 0, "pval" = 0, "n" = 0, "percent" = 0)
df_cat <- df_cat[-1]
line <- 1

cat_drop <- c()

for (factor in 2:ncol(df_init)) {
  if (length(table(df_init[,factor])) == 1) {} # drop factors wth no variation
  else if (!is.continuous(df = df_init, f = factor)) {  
    df_cat$name[line] <- colnames(df_init)[factor]  
    
    one <- df_init[df_init[,factor] == 1,]$hemoglobin_a1c_v3y0
    not_one <- df_init[df_init[,factor] != 1,]$hemoglobin_a1c_v3y0
    
    df_cat$pval[line] <- wilcox.test(one, not_one)$p.value
    cat_summary <- as.data.frame(table(df_init[,factor], useNA = "always"))
    df_cat$n[line] <- cat_summary[cat_summary$Var1==1,]$Freq[1]
    df_cat$percent[line] <- df_cat$n[line]/sum(cat_summary$Freq) * 100
      
    if (df_cat$pval[line] < 0.05) {
      signif <- c(signif, factor)
      cat_drop <- c(cat_drop, factor)
    }
    line <- line + 1
  }
}
    
df_cat <- df_cat[df_cat$name != 0,]

kable(df_cat)

print(paste0(length(cont_drop), " continuous features had a p value of less than 0.05 predicting baseline A1C with the linear model."))
print(paste0(length(cat_drop), " categorical features had a p value of less than 0.05 predicting baseline A1C with the wilcox test."))


```


# Summary of remaining features:
# Summary of remaining features:
```{r echo = FALSE, results = 'asis'}

print(paste0("total n features: ", length(unique(signif))))

# ensure that age and sex are in signif
signif <- c(signif, c(grep("age", names(df_init)), grep("sex", names(df_init))))

include_factors <- names(df_init)[unique(signif)]


cont_summary <- data.frame(matrix(nrow = 100, ncol= 4, 0))
colnames(cont_summary) <- c("name", "n", "mean", "std")

df_include <- df_init[colnames(df_init)%in%include_factors]

i <- 1
for (factor in 1:ncol(df_include)) {
  if (length(table(df_include[,factor])) > 10) {  # factor is continuous
    cont_summary$name[i] <- colnames(df_include)[factor]
    cont_summary$n[i] <- sum(!is.na(df_include[,factor]))
    cont_summary$mean[i] <- mean(df_include[,factor], na.rm = TRUE)
    cont_summary$std[i] <- sd(df_include[,factor], na.rm = TRUE)
    i <- i + 1
  } else { # factor is categorical
    cat_summary <- as.data.frame(table(df_include[,factor], useNA = "always"))
    cat_summary$Percentage <- (cat_summary$Freq*100)/total
    print(kable(cat_summary, caption = colnames(df_include)[factor]))
  }
}
cont_summary <- cont_summary[cont_summary$name != 0,]
print(kable(cont_summary, caption = "continuous variable summary"))

```





Impute missing values
``` {r echo = FALSE, message = FALSE}
# Impute missing values using k-nearest neighbors
library(DMwR)

imputed_patients <- as.integer(rownames(df_include[!complete.cases(df_include),]))

imputed_data <- knnImputation(df_include, k = 10, meth = "median", scale = F) 

# save a copy of non imputed data
non_imputed <- na.omit(df_include)


print(paste0(nrow(imputed_data)-nrow(non_imputed), " patients are dropped without imputation"))


```


```{r echo = FALSE, results = 'asis', message = FALSE}
library(car)

transform_continuous_data <- function(df, description, print_summary){

  # if lambda needed to transform data is too high and there is not a significant change in normality, do not transform the data
  summary <- data.frame("init" = c(1:length(df_include)), "name" = 0, "initial_rsquared" = 0, "transformed_rsquared" = 0, "lambda" = 0, "transformed" = 1)
  summary <- summary[-c(1)]
  line <- 1
  
  biglambda <- c()
  untransformed <- c()
  for (factor in 1:ncol(df)) {
    if (length(table(df[,factor])) > 10) {  # factor is continuous
      my_object <- df[,factor]
      name <- colnames(df)[factor]
          
      # before transformation
  
      qq_before <- qqnorm(my_object, ylab=name, main = paste0(name, " before transformation"), plot.it = FALSE)
      r_before <- cor(qq_before$x,qq_before$y)
  
      bc_dfult <- boxCox(my_object ~ 1, plotit = F, lambda = seq(-5,5,.1), eps = 0, family = "yjPower")
      bc_df <- data.frame(bc_dfult$x, bc_dfult$y)
      lambda <- bc_df[with(bc_df, order(-bc_df$bc_dfult.y)),][1,1]
          
      # after transformation
      if (lambda == 0) {transformed <- log(my_object)}
      if (lambda != 0) {transformed <- (my_object ^ lambda -1) / lambda}
            
      qq_after <- qqnorm(transformed, ylab=name, main = paste0(name, " after transformation"), plot.it = FALSE)
      r_after <- cor(qq_after$x,qq_after$y, use = "na.or.complete")
  
      if  ((r_after - r_before)/r_before < .05) {   # if there is a significant shift in r (don't care about lambda), then transform
        transformed <- my_object
        summary$transformed[line] <- 0
        untransformed <- c(untransformed, name)           # note exceptions
      } else  {
        df[,factor] <- transformed 
        if (lambda> 1) {biglambda <- c(biglambda, name)} #note big lambdas
      }
          
      summary$name[line] <- name
      summary$initial_rsquared[line] <- round(r_before*r_before, digits = 3)
      summary$transformed_rsquared[line] <- round(r_after*r_after, digits = 3)
      summary$lambda[line] <- lambda
      #summary$n_imputed[line] <- sum(is.na(df_include[,factor]))
      #summary$percent_imputed[line] <- round(sum(is.na(df_include[,factor]))/nrow(df_include), digits = 3)
      summary$minval[line] <- round(min(df[,factor], na.rm = TRUE), digits = 3)
      summary$maxval[line] <- round(max(df[,factor], na.rm = TRUE), digits = 3)
      line <- line + 1
      
      # unit scale df
      df[,factor] <- df[,factor]-min(df[,factor], na.rm = TRUE)
      df[,factor] <- df[,factor]/max(df[,factor], na.rm = TRUE)
    }
  }
  
  summary <- summary[summary$name != 0,]
  
  # note these exceptions
  if (print_summary) {
    print(kable(summary, label = description))
    print("The following continuous factors were not transformed because transformation did not significantly shift them towards normality: ")
    print(untransformed)
  }
  
  return(df)
}
```



```{r echo = FALSE, results = "asis"}
# transform the data

transformed_imputed_data <- transform_continuous_data(imputed_data, "imputed data", TRUE)
transformed_nonimputed_data <- transform_continuous_data(non_imputed, "non imputed data", FALSE)

```


  
# Compare Imputed and Non Imputed Data Sets
use a linear model to predict pcs rate
```{r, echo = FALSE, message = FALSE, results = 'asis'}

only_imputed<- transformed_imputed_data[rownames(transformed_imputed_data) %in% imputed_patients,]

imputed_lm.res <- lm(sf12_pcs_v3y0 ~ . , data = transformed_imputed_data)
print(paste0("imputed data r squared: ", summary(imputed_lm.res)$r.squared))

nonimputed_lm.res <- lm(sf12_pcs_v3y0 ~ . , data = transformed_nonimputed_data)
print(paste0("non imputed data r squared: ", summary(nonimputed_lm.res)$r.squared))

only_imputed_lm.res <- lm(sf12_pcs_v3y0 ~ . , data = only_imputed)
print(paste0("ONLY imputed data r squared: ", summary(only_imputed_lm.res)$r.squared))


```


```{r echo = FALSE, results = 'asis'}
# save selected clinical variables with patient ID and pcs rate
final_data <- transformed_imputed_data
final_data <- data.frame(final_data, egfr_data)
final_data <- final_data[c((ncol(final_data)-4):ncol(final_data), 1:(ncol(final_data)-5))]

write.csv(final_data, "../data/preprocessing_results.csv", row.names = FALSE)

```








