---
title: "Reference Models"
output: word_document
---

Load data
```{r Load Data, echo=FALSE, message=FALSE}
#untar <- read.csv("../data/zscore_untargeted.csv")
clin <- read.csv("../data/partitioning.csv")
names(clin)[1] <- "patients"
#untarmet <- read.csv("../data/CRIC_clinical_creatinine_normalized_samplemetadata.csv", header = FALSE)

#jing <- read.csv("../data/60 var selected by 8 models from lasso RF TB (models with both ions and cli).csv")
#jing_ions <- as.numeric(unique(substring(jing$Ion, 5,15)))

# save copy of original untargeted ions 
#original_untar <- untar

# only include V3Y0
#untar <- untar[untar$visit == "V3Y0",]

#pats <- untar$patientid

# include only jing ions
#untar <- untar[,names(untar) %in% paste0("V", jing_ions)]

# combine untargeted and clinical data
#untar$patients <- pats
#combined <- merge(untar, clin, by = "patients")


# separate into groups, remove patient information
#train1 <- combined[combined$group == 0,][-1]
#train2 <- combined[combined$group == 1,][-1]
#train3 <- combined[combined$group == 2,][-1]

# metadata <- c("group", "rawdata.mu_egfr", "rawdata.sigma_egfr", "rawdata.r_group", "rawdata.v_group")

#write.csv(train1, "../data/clin_and_untar_train1.csv", row.names = FALSE)
#write.csv(train2, "../data/clin_and_untar_train2.csv", row.names = FALSE)
#write.csv(train3, "../data/clin_and_untar_train3.csv", row.names = FALSE)

#clin1 <- clin[clin$group == 0,][-1]
#clin2 <- clin[clin$group == 1,][-1]
#clin3 <- clin[clin$group == 2,][-1]

#write.csv(clin1, "../data/clin_train1.csv", row.names = FALSE)
#write.csv(clin2, "../data/clin_train2.csv", row.names = FALSE)
#write.csv(clin3, "../data/clin_train3.csv", row.names = FALSE)

#train1 <- read.csv("../data/clin_and_untar_train1.csv")
#train2 <- read.csv("../data/clin_and_untar_train2.csv")
#train3 <- read.csv("../data/clin_and_untar_train3.csv")

#clin1 <- read.csv("../data/clin_train1.csv")
#clin2 <- read.csv("../data/clin_train2.csv")
#clin3 <- read.csv("../data/clin_train3.csv")

```


```{r echo=FALSE, message=FALSE}
library(caret)
library(ModelMetrics)
library(randomForest)
library(knitr)
library(caret)
library(e1071)
library(pROC)
library(ROCR)

linear_model <- function(model, data, newdata, predict) {
  if (predict == "rate") {
    training_MSE <- mse(model$fitted.values, data$rawdata.mu_egfr)
    pred <- predict(model, newdata = newdata)
    testing_MSE <- mse(pred,newdata$rawdata.mu_egfr)
    se <- (pred-newdata$rawdata.mu_egfr)^2
  } else if (predict == "volatility") {
    training_MSE <- mse(model$fitted.values, data$rawdata.sigma_egfr)
    pred <- predict(model, newdata = newdata)
    testing_MSE <- mse(pred,newdata$rawdata.sigma_egfr)
    se <- (pred-newdata$rawdata.sigma_egfr)^2
  }
  return(se)
}


rf_model <- function(model, data, newdata, predict) {
  if (predict == "rate") {
    training_MSE <- mse(model$predicted, data$rawdata.mu_egfr)
    pred <- predict(model, newdata = newdata)
    testing_MSE <- mse(pred,newdata$rawdata.mu_egfr)
    #t <- roc(predictor = as.numeric(model$predicted), as.factor(as.numeric(data$rawdata.r_group <= 2)), plot = TRUE, ci = TRUE)
    #print(paste0("training MSE: ", training_MSE))
    #t$auc
    #t$ci
    #v <- roc(predictor = as.numeric(pred), as.factor(as.numeric(newdata$rawdata.r_group <= 2)), plot = TRUE, ci = TRUE)
    #print(paste0("testing MSE: ", testing_MSE))
    #v$auc
    #v$ci
    se <- (pred-newdata$rawdata.mu_egfr)^2
  } else if (predict == "volatility") {
    training_MSE <- mse(model$predicted, data$rawdata.sigma_egfr)
    pred <- predict(model, newdata = newdata)
    testing_MSE <- mse(pred,newdata$rawdata.sigma_egfr)
    #t <- roc(predictor = as.numeric(model$predicted), as.factor(as.numeric(data$rawdata.v_group <= 2)), plot = TRUE, ci = TRUE)
    #print(paste0("training MSE: ", training_MSE))
    #t$auc
    #t$ci
    #v <- roc(predictor = as.numeric(pred), as.factor(as.numeric(newdata$rawdata.v_group <= 2)), plot = TRUE, ci = TRUE)
    #print(paste0("testing MSE: ", testing_MSE))
    #v$auc
    #v$ci
    se <- (pred-newdata$rawdata.sigma_egfr)^2
  }
  return(se)
}


```



#Linear Models
Predicting EGFR Rate with Clinical Variables
```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(clin1, clin2) )
lcr1 <- linear_model(model, rbind(clin1,clin2), clin3, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(clin2, clin3))
lcr2 <- linear_model(model, rbind(clin2,clin3), clin1, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(clin1, clin3))
lcr3 <- linear_model(model, rbind(clin1,clin3), clin2, predict = "rate")

```

Predicting EGFR Volatility with Clinical Variables
```{r echo = FALSE, message = FALSE, results = 'asis', warning = FALSE}
model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(clin1, clin2))
lcv1 <- linear_model(model, rbind(clin1,clin2), clin3, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(clin2, clin3))
lcv2 <- linear_model(model, rbind(clin2,clin3), clin1, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(clin1, clin3))
lcv3 <- linear_model(model, rbind(clin1,clin3), clin2, predict = "volatility")

```

Predicting EGFR Rate with Clinical Variables and Untargeted Ions
```{r echo = FALSE, message = FALSE, results = 'asis', warning = FALSE}
model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(train1, train2))
lbr1 <- linear_model(model, rbind(train1,train2), train3, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(train2, train3))
lbr2 <- linear_model(model, rbind(train2,train3), train1, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(train1, train3))
lbr3 <- linear_model(model, rbind(train1,train3), train2, predict = "rate")

```

Predicting EGFR Volatility with combical Variables and Untargeted Ions
```{r echo = FALSE, message = FALSE, results = 'asis', warning = FALSE}
model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(train1, train2))
lbv1 <- linear_model(model, rbind(train1,train2), train3, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(train2, train3))
lbv2 <- linear_model(model, rbind(train2,train3), train1, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(train1, train3))
lbv3 <- linear_model(model, rbind(train1,train3), train2, predict = "volatility")

```

#Random Forest
Predicting EGFR Rate with Clinical Variables
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(clin1,clin2), importance = TRUE, na.action = "na.exclude")
rfcr1 <- rf_model(rf, rbind(clin1,clin2), clin3, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(clin2,clin3), importance = TRUE, na.action = "na.exclude")
rfcr2 <- rf_model(rf, rbind(clin2,clin3), clin1, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(clin1,clin3), importance = TRUE, na.action = "na.exclude")
rfcr3 <- rf_model(rf, rbind(clin1,clin3), clin2, predict = "rate")


```

Predicting EGFR Volatility with Clinical Variables
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(clin1,clin2), importance = TRUE, na.action = "na.exclude")
rfcv1 <- rf_model(rf, rbind(clin1,clin2), clin3, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(clin2,clin3), importance = TRUE, na.action = "na.exclude")
rfcv2 <- rf_model(rf, rbind(clin2,clin3), clin1, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(clin1,clin3), importance = TRUE, na.action = "na.exclude")
rfcv3 <- rf_model(rf, rbind(clin1,clin3), clin2, predict = "volatility")

```

Predicting EGFR Rate with Clinical Variables and Untargeted Ions
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(train1,train2), importance = TRUE, na.action = "na.exclude")
rfbr1 <- rf_model(rf, rbind(train1,train2), train3, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(train2,train3), importance = TRUE, na.action = "na.exclude")
rfbr2 <- rf_model(rf, rbind(train2,train3), train1, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(train1,train3), importance = TRUE, na.action = "na.exclude")
rfbr3 <- rf_model(rf, rbind(train1,train3), train2, predict = "rate")

```

Predicting EGFR Volatility with clinical Variables and Untargeted Ions
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(train1,train2), importance = TRUE, na.action = "na.exclude")
rfbv1 <- rf_model(rf, rbind(train1,train2), train3, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(train2,train3), importance = TRUE, na.action = "na.exclude")
rfbv2 <- rf_model(rf, rbind(train2,train3), train1, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(train1,train3), importance = TRUE, na.action = "na.exclude")
rfbv3 <- rf_model(rf, rbind(train1,train3), train2, predict = "volatility")

```


# Save Results
```{r echo = FALSE, message = FALSE}
df <- data.frame("lbr1" = lbr1,
                 "lbr2" = c(lbr2, rep(NA, 2)),
                 "lbr3" = lbr3,
                 "lbv1" = lbv1,
                 "lbv2" = c(lbv2, rep(NA, 2)),
                 "lbv3" = lbv3,
                 "rfbr1" = rfbr1,
                 "rfbr2" = c(rfbr2, rep(NA, 2)),
                 "rfbr3" = rfbr3,
                 "rfbv1" = rfbv1,
                 "rfbv2" = c(rfbv2, rep(NA, 2)),
                 "rfbv3" = rfbv3
                 )

write.csv(df, "../data/reference_model_df.csv")
```


# Plot Results
Plots for each model
```{r echo = FALSE, message = FALSE}
boxplot(lcr1, lcr2, lcr3, outline = FALSE)
title("Linear model for Rate with Clinical, 3x Cross Val.")

boxplot(lcv1, lcv2, lcv3, outline = FALSE)
title("Linear model for Volatility with Clinical, 3x Cross Val.")

boxplot(lbr1, lbr2, lbr3, outline = FALSE)
title("Linear model for Rate with Clinical and Ions, 3x Cross Val.")

boxplot(lbv1, lbv2, lbv3, outline = FALSE)
title("Linear model for Volatility with Clinical and Ions, 3x Cross Val.")

boxplot(rfcr1, rfcr2, rfcr3, outline = FALSE)
title("Random Forest model for Rate with Clinical, 3x Cross Val.")

boxplot(rfcv1, rfcv2, rfcv3, outline = FALSE)
title("Random Forest model for Volatility with Clinical, 3x Cross Val.")

boxplot(rfbr1, rfbr2, rfbr3, outline = FALSE)
title("Random Forest model for Rate with Clinical and Ions, 3x Cross Val.")

boxplot(rfbv1, rfbv2, rfbv3, outline = FALSE)
title("Random Forest model for Volatility with Clinical and Ions, 3x Cross Val.")

```



Plots comparing models (using ggplot and with p vals comparing means, first rep from cross val)
```{r echo = FALSE, message = FALSE}
library(ggplot2)
library(ggpubr)
rate_data <- data.frame("model" = c(rep("Linear: Clin", length(lcr1)),
                                    rep("Linear: Clin and Ions", length(lbr1)),
                                    rep("RF: Clin", length(rfcr1)),
                                    rep("RF: Clin and Ions", length(rfbr1))),
                        "error" = c(lcr1, lbr1, rfcr1, rfbr1)
)

rate_data$model <- factor(rate_data$model, levels=c("Linear: Clin", "Linear: Clin and Ions", "RF: Clin", "RF: Clin and Ions"))

compare_means(error~model, data = rate_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and Ions"), c("RF: Clin", "RF: Clin and Ions"))

ggplot(rate_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.017, tip.length = 0.001) + 
  ylim(0, 0.018)+
  ggtitle("MSE for Models Predicting EGFR Rate of Decline")


vol_data <- data.frame("model" = c(rep("Linear: Clin", length(lcv1)),
                                   rep("Linear: Clin and Ions", length(lbv1)),
                                   rep("RF: Clin", length(rfcv1)),
                                   rep("RF: Clin and Ions", length(rfbv1))),
                       "error" = c(lcv1, lbv1, rfcv1, rfbv1)
)

vol_data$model <- factor(vol_data$model, levels=c("Linear: Clin", "Linear: Clin and Ions", "RF: Clin", "RF: Clin and Ions"))

compare_means(error~model, data = vol_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and Ions"), c("RF: Clin", "RF: Clin and Ions"))

ggplot(vol_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.0055, tip.length = 0.001) + 
  ylim(0, 0.006)+
  ggtitle("MSE for Models Predicting EGFR Volatility")

```

Table of MSEs (first rep from cross val)
```{r echo = FALSE, message = FALSE}
mse <- data.frame("Linear with Clinical" = c(round(mean(lcr1), digits = 3), round(mean(lcv1), digits = 3)),
                  "Linear with Clinical and Ions" = c(round(mean(lbr1), digits = 3), round(mean(lbv1), digits = 3)),
                  "RF with Clinical" = c(round(mean(rfcr1), digits = 3), round(mean(rfcv1), digits = 3)),
                  "RF with Clinical and Ions" = c(round(mean(rfbr1), digits = 3), round(mean(rfbv1), digits = 3)))
rownames(mse) <- c("Rate", "Volatility")
kable(mse)

write.csv(mse, "../data/reference_model_mses.csv")
```


## Three fold CV using coordinates

# PCA
Load coordinate data
```{r echo = FALSE, message = FALSE}
pca_coords <- read.csv("../data/pca_coords.csv")

pca_coords_t1 <- merge(clin[clin$group == 0,], pca_coords, by = "patients")[-1]
pca_coords_t2 <- merge(clin[clin$group == 1,], pca_coords, by = "patients")[-1]
pca_coords_t3 <- merge(clin[clin$group == 2,], pca_coords, by = "patients")[-1]

```

Linear Models
Predicting Rate
```{r echo = FALSE, message = FALSE}
model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(pca_coords_t1, pca_coords_t2))
pca_lbr1 <- linear_model(model, rbind(pca_coords_t1,pca_coords_t2), pca_coords_t3, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(pca_coords_t2, pca_coords_t3))
pca_lbr2 <- linear_model(model, rbind(pca_coords_t2,pca_coords_t3), pca_coords_t1, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(pca_coords_t1, pca_coords_t3))
pca_lbr3 <- linear_model(model, rbind(pca_coords_t1,pca_coords_t3), pca_coords_t2, predict = "rate")

```

Predicting Volatility
```{r echo = FALSE, message = FALSE}
model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(pca_coords_t1, pca_coords_t2))
pca_lbv1 <- linear_model(model, rbind(pca_coords_t1,pca_coords_t2), pca_coords_t3, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(pca_coords_t2, pca_coords_t3))
pca_lbv2 <- linear_model(model, rbind(pca_coords_t2,pca_coords_t3), pca_coords_t1, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(pca_coords_t1, pca_coords_t3))
pca_lbv3 <- linear_model(model, rbind(pca_coords_t1,pca_coords_t3), pca_coords_t2, predict = "volatility")

```


Random Forest
Predicting Rate
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(pca_coords_t1,pca_coords_t2), importance = TRUE, na.action = "na.exclude")
pca_rfbr1 <- rf_model(rf, rbind(pca_coords_t1,pca_coords_t2), pca_coords_t3, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(pca_coords_t2,pca_coords_t3), importance = TRUE, na.action = "na.exclude")
pca_rfbr2 <- rf_model(rf, rbind(pca_coords_t2,pca_coords_t3), pca_coords_t1, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(pca_coords_t1,pca_coords_t3), importance = TRUE, na.action = "na.exclude")
pca_rfbr3 <- rf_model(rf, rbind(pca_coords_t1,pca_coords_t3), pca_coords_t2, predict = "rate")

```

Predicting Volatility
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(pca_coords_t1,pca_coords_t2), importance = TRUE, na.action = "na.exclude")
pca_rfbv1 <- rf_model(rf, rbind(pca_coords_t1,pca_coords_t2), pca_coords_t3, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(pca_coords_t2,pca_coords_t3), importance = TRUE, na.action = "na.exclude")
pca_rfbv2 <- rf_model(rf, rbind(pca_coords_t2,pca_coords_t3), pca_coords_t1, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(pca_coords_t1,pca_coords_t3), importance = TRUE, na.action = "na.exclude")
pca_rfbv3 <- rf_model(rf, rbind(pca_coords_t1,pca_coords_t3), pca_coords_t2, predict = "volatility")

```

Plots comparing models 
```{r echo = FALSE, message = FALSE}

pca_rate_data <- data.frame("model" = c(rep("Linear: Clin", length(lcr1)),
                                    rep("Linear: Clin and PCA Coords", length(pca_lbr1)),
                                    rep("RF: Clin", length(rfcr1)),
                                    rep("RF: Clin and PCA Coords", length(pca_rfbr1))),
                        "error" = c(lcr1, pca_lbr1, rfcr1, pca_rfbr1)
)

pca_rate_data$model <- factor(pca_rate_data$model, levels=c("Linear: Clin", "Linear: Clin and PCA Coords", "RF: Clin", "RF: Clin and PCA Coords"))

compare_means(error~model, data = pca_rate_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and PCA Coords"), c("RF: Clin", "RF: Clin and PCA Coords"))

ggplot(pca_rate_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.017, tip.length = 0.001) + 
  ylim(0, 0.018)+
  ggtitle("MSE for Models Predicting EGFR Rate of Decline")


pca_vol_data <- data.frame("model" = c(rep("Linear: Clin", length(lcv1)),
                                   rep("Linear: Clin and PCA Coords", length(pca_lbv1)),
                                   rep("RF: Clin", length(rfcv1)),
                                   rep("RF: Clin and PCA Coords", length(pca_rfbv1))),
                       "error" = c(lcv1, pca_lbv1, rfcv1, pca_rfbv1)
)

pca_vol_data$model <- factor(pca_vol_data$model, levels=c("Linear: Clin", "Linear: Clin and PCA Coords", "RF: Clin", "RF: Clin and PCA Coords"))

compare_means(error~model, data = pca_vol_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and PCA Coords"), c("RF: Clin", "RF: Clin and PCA Coords"))

ggplot(pca_vol_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.0055, tip.length = 0.001) + 
  ylim(0, 0.006)+
  ggtitle("MSE for Models Predicting EGFR Volatility")

```

# tSNE
Load coordinate data
```{r echo = FALSE, message = FALSE}
tsne_coords <- read.csv("../data/tsne_coords.csv")

tsne_coords_t1 <- merge(clin[clin$group == 0,], tsne_coords, by = "patients")[-1]
tsne_coords_t2 <- merge(clin[clin$group == 1,], tsne_coords, by = "patients")[-1]
tsne_coords_t3 <- merge(clin[clin$group == 2,], tsne_coords, by = "patients")[-1]

```

Linear Models
Predicting Rate
```{r echo = FALSE, message = FALSE}
model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(tsne_coords_t1, tsne_coords_t2))
tsne_lbr1 <- linear_model(model, rbind(tsne_coords_t1,tsne_coords_t2), tsne_coords_t3, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(tsne_coords_t2, tsne_coords_t3))
tsne_lbr2 <- linear_model(model, rbind(tsne_coords_t2,tsne_coords_t3), tsne_coords_t1, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(tsne_coords_t1, tsne_coords_t3))
tsne_lbr3 <- linear_model(model, rbind(tsne_coords_t1,tsne_coords_t3), tsne_coords_t2, predict = "rate")

```

Predicting Volatility
```{r echo = FALSE, message = FALSE}
model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(tsne_coords_t1, tsne_coords_t2))
tsne_lbv1 <- linear_model(model, rbind(tsne_coords_t1,tsne_coords_t2), tsne_coords_t3, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(tsne_coords_t2, tsne_coords_t3))
tsne_lbv2 <- linear_model(model, rbind(tsne_coords_t2,tsne_coords_t3), tsne_coords_t1, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(tsne_coords_t1, tsne_coords_t3))
tsne_lbv3 <- linear_model(model, rbind(tsne_coords_t1,tsne_coords_t3), tsne_coords_t2, predict = "volatility")

```


Random Forest
Predicting Rate
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(tsne_coords_t1,tsne_coords_t2), importance = TRUE, na.action = "na.exclude")
tsne_rfbr1 <- rf_model(rf, rbind(tsne_coords_t1,tsne_coords_t2), tsne_coords_t3, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(tsne_coords_t2,tsne_coords_t3), importance = TRUE, na.action = "na.exclude")
tsne_rfbr2 <- rf_model(rf, rbind(tsne_coords_t2,tsne_coords_t3), tsne_coords_t1, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(tsne_coords_t1,tsne_coords_t3), importance = TRUE, na.action = "na.exclude")
tsne_rfbr3 <- rf_model(rf, rbind(tsne_coords_t1,tsne_coords_t3), tsne_coords_t2, predict = "rate")

```

Predicting Volatility
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(tsne_coords_t1,tsne_coords_t2), importance = TRUE, na.action = "na.exclude")
tsne_rfbv1 <- rf_model(rf, rbind(tsne_coords_t1,tsne_coords_t2), tsne_coords_t3, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(tsne_coords_t2,tsne_coords_t3), importance = TRUE, na.action = "na.exclude")
tsne_rfbv2 <- rf_model(rf, rbind(tsne_coords_t2,tsne_coords_t3), tsne_coords_t1, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(tsne_coords_t1,tsne_coords_t3), importance = TRUE, na.action = "na.exclude")
tsne_rfbv3 <- rf_model(rf, rbind(tsne_coords_t1,tsne_coords_t3), tsne_coords_t2, predict = "volatility")

```

Plots comparing models 
```{r echo = FALSE, message = FALSE}

tsne_rate_data <- data.frame("model" = c(rep("Linear: Clin", length(lcr1)),
                                    rep("Linear: Clin and TSNE Coords", length(tsne_lbr1)),
                                    rep("RF: Clin", length(rfcr1)),
                                    rep("RF: Clin and TSNE Coords", length(tsne_rfbr1))),
                        "error" = c(lcr1, tsne_lbr1, rfcr1, tsne_rfbr1)
)

tsne_rate_data$model <- factor(tsne_rate_data$model, levels=c("Linear: Clin", "Linear: Clin and TSNE Coords", "RF: Clin", "RF: Clin and TSNE Coords"))

compare_means(error~model, data = tsne_rate_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and TSNE Coords"), c("RF: Clin", "RF: Clin and TSNE Coords"))

ggplot(tsne_rate_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.017, tip.length = 0.001) + 
  ylim(0, 0.018)+
  ggtitle("MSE for Models Predicting EGFR Rate of Decline")


tsne_vol_data <- data.frame("model" = c(rep("Linear: Clin", length(lcv1)),
                                   rep("Linear: Clin and TSNE Coords", length(tsne_lbv1)),
                                   rep("RF: Clin", length(rfcv1)),
                                   rep("RF: Clin and TSNE Coords", length(tsne_rfbv1))),
                       "error" = c(lcv1, tsne_lbv1, rfcv1, tsne_rfbv1)
)

tsne_vol_data$model <- factor(tsne_vol_data$model, levels=c("Linear: Clin", "Linear: Clin and TSNE Coords", "RF: Clin", "RF: Clin and TSNE Coords"))

compare_means(error~model, data = tsne_vol_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and TSNE Coords"), c("RF: Clin", "RF: Clin and TSNE Coords"))

ggplot(tsne_vol_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.0055, tip.length = 0.001) + 
  ylim(0, 0.006)+
  ggtitle("MSE for Models Predicting EGFR Volatility")

```


# SOM
Load coordinate data
```{r echo = FALSE, message = FALSE}
som_coords <- read.csv("../data/som_coords.csv")

som_coords_t1 <- merge(clin[clin$group == 0,], som_coords, by = "patients")[-1]
som_coords_t2 <- merge(clin[clin$group == 1,], som_coords, by = "patients")[-1]
som_coords_t3 <- merge(clin[clin$group == 2,], som_coords, by = "patients")[-1]

```

Linear Models
Predicting Rate
```{r echo = FALSE, message = FALSE}
model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(som_coords_t1, som_coords_t2))
som_lbr1 <- linear_model(model, rbind(som_coords_t1,som_coords_t2), som_coords_t3, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(som_coords_t2, som_coords_t3))
som_lbr2 <- linear_model(model, rbind(som_coords_t2,som_coords_t3), som_coords_t1, predict = "rate")

model <- lm(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(som_coords_t1, som_coords_t3))
som_lbr3 <- linear_model(model, rbind(som_coords_t1,som_coords_t3), som_coords_t2, predict = "rate")

```

Predicting Volatility
```{r echo = FALSE, message = FALSE}
model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(som_coords_t1, som_coords_t2))
som_lbv1 <- linear_model(model, rbind(som_coords_t1,som_coords_t2), som_coords_t3, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(som_coords_t2, som_coords_t3))
som_lbv2 <- linear_model(model, rbind(som_coords_t2,som_coords_t3), som_coords_t1, predict = "volatility")

model <- lm(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group - group, data = rbind(som_coords_t1, som_coords_t3))
som_lbv3 <- linear_model(model, rbind(som_coords_t1,som_coords_t3), som_coords_t2, predict = "volatility")

```


Random Forest
Predicting Rate
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(som_coords_t1,som_coords_t2), importance = TRUE, na.action = "na.exclude")
som_rfbr1 <- rf_model(rf, rbind(som_coords_t1,som_coords_t2), som_coords_t3, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(som_coords_t2,som_coords_t3), importance = TRUE, na.action = "na.exclude")
som_rfbr2 <- rf_model(rf, rbind(som_coords_t2,som_coords_t3), som_coords_t1, predict = "rate")

rf <- randomForest(rawdata.mu_egfr ~ . - rawdata.sigma_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(som_coords_t1,som_coords_t3), importance = TRUE, na.action = "na.exclude")
som_rfbr3 <- rf_model(rf, rbind(som_coords_t1,som_coords_t3), som_coords_t2, predict = "rate")

```

Predicting Volatility
```{r echo = FALSE, message = FALSE}
rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(som_coords_t1,som_coords_t2), importance = TRUE, na.action = "na.exclude")
som_rfbv1 <- rf_model(rf, rbind(som_coords_t1,som_coords_t2), som_coords_t3, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(som_coords_t2,som_coords_t3), importance = TRUE, na.action = "na.exclude")
som_rfbv2 <- rf_model(rf, rbind(som_coords_t2,som_coords_t3), som_coords_t1, predict = "volatility")

rf <- randomForest(rawdata.sigma_egfr ~ . - rawdata.mu_egfr - rawdata.r_group - rawdata.v_group -group, data = rbind(som_coords_t1,som_coords_t3), importance = TRUE, na.action = "na.exclude")
som_rfbv3 <- rf_model(rf, rbind(som_coords_t1,som_coords_t3), som_coords_t2, predict = "volatility")

```

Plots comparing models 
```{r echo = FALSE, message = FALSE}

som_rate_data <- data.frame("model" = c(rep("Linear: Clin", length(lcr1)),
                                    rep("Linear: Clin and SOM Coords", length(som_lbr1)),
                                    rep("RF: Clin", length(rfcr1)),
                                    rep("RF: Clin and SOM Coords", length(som_rfbr1))),
                        "error" = c(lcr1, som_lbr1, rfcr1, som_rfbr1)
)

som_rate_data$model <- factor(som_rate_data$model, levels=c("Linear: Clin", "Linear: Clin and SOM Coords", "RF: Clin", "RF: Clin and SOM Coords"))

compare_means(error~model, data = som_rate_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and SOM Coords"), c("RF: Clin", "RF: Clin and SOM Coords"))

ggplot(som_rate_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.017, tip.length = 0.001) + 
  ylim(0, 0.018)+
  ggtitle("MSE for Models Predicting EGFR Rate of Decline")


som_vol_data <- data.frame("model" = c(rep("Linear: Clin", length(lcv1)),
                                   rep("Linear: Clin and SOM Coords", length(som_lbv1)),
                                   rep("RF: Clin", length(rfcv1)),
                                   rep("RF: Clin and SOM Coords", length(som_rfbv1))),
                       "error" = c(lcv1, som_lbv1, rfcv1, som_rfbv1)
)

som_vol_data$model <- factor(som_vol_data$model, levels=c("Linear: Clin", "Linear: Clin and SOM Coords", "RF: Clin", "RF: Clin and SOM Coords"))

compare_means(error~model, data = som_vol_data)
my_comparisons <- list(c("Linear: Clin", "Linear: Clin and SOM Coords"), c("RF: Clin", "RF: Clin and SOM Coords"))

ggplot(som_vol_data, aes(x = model, y = error))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", label.y = 0.0055, tip.length = 0.001) + 
  ylim(0, 0.006)+
  ggtitle("MSE for Models Predicting EGFR Volatility")

```















